---
title: "Course overview"
author: "Dr. Olanrewaju Michael Akande"
date: "Module 0.1"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

class: center, middle

# Welcome to STA 360 and STA 602L!



---

## What is this course about?

<i class="fa fa-book fa-2x"></i> Learn the foundations of Bayesian inference.

--

<i class="fa fa-folder-open fa-2x"></i> Work through the theory of several Bayesian models.

--

<i class="fa fa-tasks fa-2x"></i> Use Bayesian models to answer inferential questions.

--

<i class="fa fa-database fa-2x"></i> Apply the models to several different problem sets.

--

<i class="fa fa-refresh fa-spin fa-2x"></i> ".hlight[Prior] $\rightarrow$ .hlight[likelihood] $\rightarrow$ .hlight[posterior]" over and over again!

--

<i class="fa fa-clock fa-2x"></i> We will follow the Hoff book closely -- roughly one chapter per week.

--

--- 
<i class="fa fa-quote-left fa-2x fa-pull-left fa-border" aria-hidden="true"></i>
<i class="fa fa-quote-right fa-2x fa-pull-right fa-border" aria-hidden="true"></i>
A Bayesian version will usually make things better...

-- Andrew Gelman.



---

## Instructor

[Dr. Olanrewaju Michael Akande](https://akandelanre.github.io.)

<i class="fa fa-envelope"></i> &nbsp; [olanrewaju.akande@duke.edu](mailto:olanrewaju.akande@duke.edu) <br>
<i class="fa fa-home"></i> &nbsp; [akandelanre.github.io.](https://akandelanre.github.io/IDS702_F19/) <br>
<i class="fa fa-university"></i> &nbsp; [256 Gross Hall](https://maps.duke.edu/?id=21#!m/6866?s/Gross%20Hall) <br>
<i class="fa fa-calendar"></i> &nbsp; Wed 9:00 - 10:00am; Thur 11:45 - 12:45pm (still subject to change!)



---

## Lead TA

[Jordan Bryan](https://stat.duke.edu/people/jordan-bryan) (mainly for STA 360)

<i class="fa fa-envelope"></i> &nbsp; [jordan.bryan@duke.edu](mailto:jordan.bryan@duke.edu) <br>


## TAs

[Bai Li](https://stat.duke.edu/people/bai-li)

<i class="fa fa-envelope"></i> &nbsp; [bai.li@duke.edu](mailto:bai.li@duke.edu) <br>
<i class="fa fa-calendar"></i> &nbsp; Wed 3:00 - 5:00pm <br>
<i class="fa fa-university"></i> &nbsp; [Old Chem 025](https://maps.duke.edu/map/?id=21#!s/key=Old%20Chemistry?m/2766)

[Zhuoqun (Carol) Wang](https://stat.duke.edu/people/zhuoqun-wang-0)

<i class="fa fa-envelope"></i> &nbsp; [zhuoqun.wang@duke.edu](mailto:zhuoqun.wang@duke.edu) <br>
<i class="fa fa-calendar"></i> &nbsp; Tues 3:00 - 5:00pm <br>
<i class="fa fa-university"></i> &nbsp; [Old Chem 025](https://maps.duke.edu/map/?id=21#!s/key=Old%20Chemistry?m/2766)



---

## FAQs

All materials and information will be posted on the course webpage:

[https://sta-602l-s20.github.io/Course-Website/](https://sta-602l-s20.github.io/Course-Website/)

--

- How much theory will this class cover? A lot! Make sure you are especially comfortable working with probability distributions.

--

- Am I prepared to take this course? Yes, if you are familiar with the topics covered in the course prerequisites.

--

- Will we be doing "very heavy" computing? Not too heavy but yes, a good amount. You will be expected to be able to write your own MCMC sampler later on.

--

- What computing language will we use? R!

--

- What if I don't know R? This course assumes you already know R but you can still learn on the fly (you must be self-motivated). Here are some resources for you: [https://sta-602l-s20.github.io/Course-Website/resources/](https://sta-602l-s20.github.io/Course-Website/resources/).


---

class: center, middle

# Course structure and policies


---
## Course structure and policies

- All on the website (here: [https://sta-602l-s20.github.io/Course-Website/policies/](https://sta-602l-s20.github.io/Course-Website/policies/))

--

- Make use of the teaching team's office hours, we're here to help!

--

- Do not hesitate to come to my office during office hours or by appointment to discuss a homework problem or any aspect of the course. 

--

- When the teaching team has announcements for you we will send an email to your Duke email address. Please make sure to check your email daily.

--

- Please refrain from texting or using your computer for anything other than coursework during class.


---
## Other details

- What topics will we cover? Refer to Section 9 of the syllabus (here: https://sta-602l-s20.github.io/Course-Website/syllabus_pdf/Syllabus.pdf).

--

- If you are auditing this course, remember to complete the audit form for the graduate school.

--

- Confirm that you have access to Sakai and Gradescope.


