---
title: "STA 360/602L: Module 0.1"
subtitle: "Course overview"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

class: center, middle

# Welcome to STA 360/602L!



---

## What is this course about?

<i class="fa fa-book fa-2x"></i> Learn the foundations of Bayesian inference.

--

<i class="fa fa-folder-open fa-2x"></i> Work through the theory of several Bayesian models.

--

<i class="fa fa-tasks fa-2x"></i> Use Bayesian models to answer inferential questions.

--

<i class="fa fa-database fa-2x"></i> Apply the models to several different problem sets.

--

<i class="fa fa-refresh fa-spin fa-2x"></i> ".hlight[Prior] $\rightarrow$ .hlight[likelihood] $\rightarrow$ .hlight[posterior]" over and over again!

--

<i class="fa fa-clock fa-2x"></i> We will follow the Hoff book closely -- roughly two chapters per week.

--

--- 
<i class="fa fa-quote-left fa-2x fa-pull-left fa-border" aria-hidden="true"></i>
<i class="fa fa-quote-right fa-2x fa-pull-right fa-border" aria-hidden="true"></i>
A Bayesian version will usually make things better...

-- Andrew Gelman.



---
## Instructor

[Dr. Olanrewaju Michael Akande](https://akandelanre.github.io.)

<i class="fa fa-envelope"></i> &nbsp; [olanrewaju.akande@duke.edu](mailto:olanrewaju.akande@duke.edu) <br>
<i class="fa fa-home"></i> &nbsp; [akandelanre.github.io.](https://akandelanre.github.io/IDS702_F19/) <br>
<i class="fa fa-calendar"></i> &nbsp; See course website <br>
<i class="fa fa-university"></i> &nbsp; See course website


---
## TAs

[Christine Shen](https://stat.duke.edu/people/christine-shen)

<i class="fa fa-envelope"></i> &nbsp; [yueming.shen@duke.edu](mailto:yueming.shen@duke.edu) <br>
<i class="fa fa-calendar"></i> &nbsp; See course website <br>
<i class="fa fa-university"></i> &nbsp; See course website <br>

[Bo Liu](https://stat.duke.edu/people/bo-liu-0)

<i class="fa fa-envelope"></i> &nbsp; [bo.liu1997@duke.edu](mailto:bo.liu1997@duke.edu) <br>
<i class="fa fa-calendar"></i> &nbsp; See course website <br>
<i class="fa fa-university"></i> &nbsp; See course website



---

## FAQs

All materials and information will be posted on the course webpage:

[https://sta-360-602l-su20.github.io/Course-Website/](https://sta-360-602l-su20.github.io/Course-Website/)

--

- How much theory will this class cover? A lot! Make sure you are especially comfortable working with probability distributions.

--

- Am I prepared to take this course? Yes, if you are familiar with the topics covered in the course prerequisites.

--

- Will we be doing "very heavy" computing? Not too heavy but yes, a good amount. You will be expected to be able to write your own MCMC sampler later on.

--

- What computing language will we use? R!

--

- What if I don't know R? This course assumes you already know R but you can still learn on the fly (you must be self-motivated). Here are some resources for you: [https://sta-360-602l-su20.github.io/Course-Website/resources/](https://sta-360-602l-su20.github.io/Course-Website/resources/).



---

## FAQs

- What if I can't remember the topics in the prerequisites?
  1. Review Chapters 1 to 5 of the [Casella and Berger book](https://mybiostats.files.wordpress.com/2015/03/casella-berger.pdf)
  
  2. You can find the solution manual [here](http://www.ams.sunysb.edu/~zhu/ams570/Solutions-Casella-Berger.pdf)
  
  3. Focus on the following topics:
      + basic probability theory, random variables, transformations of random variables, expectations of random variables, common families of probability distribution functions including multivariate distributions
      + concepts of convergence, principles of statistical inference, likelihood based inference, sampling distributions and hypothesis testing.


---

class: center, middle

# Course structure and policies


---
## Course structure and policies

- All on the website (here: [https://sta-360-602l-su20.github.io/Course-Website/policies/](https://sta-360-602l-su20.github.io/Course-Website/policies/))

--

- Make use of the teaching team's office hours, we're here to help!

--

- Do not hesitate to come to my online office hours or you can also make an appointment to discuss a homework problem or any aspect of the course. 

--

- When the teaching team has announcements for you we will send an email to your Duke email address. Please make sure to check your email daily.

--

- Try as much as possible to refrain from texting or using your computer for anything other than coursework while watching the lecture videos and during discussion sessions.


---
## Other details

- What topics will we cover? Refer to Section 11 of the syllabus (here: [STA 360 Syllabus](https://sta-360-602l-su20.github.io/Course-Website/syllabus_360_pdf/Syllabus_360.pdf) or [STA 602L Syllabus](https://sta-360-602l-su20.github.io/Course-Website/syllabus_602_pdf/Syllabus_602.pdf)).

--

- Also refer to the schedule on the website for updated breakdown of the courses. Remember to refresh the page frequently. See here: [Class Schedule](https://sta-360-602l-su20.github.io/Course-Website/syllabus/).

--

- If you are auditing this course, remember to complete the necessary audit forms.

--

- Confirm that you have access to Sakai, Piazza and Gradescope.



---
## Other details

Finally, here are some readings to keep you busy. Make sure to glance through them within the next week or so.
  1. Efron, B., 1986. Why isn't everyone a Bayesian?. The American Statistician, 40(1), pp. 1-5.
  2. Gelman, A., 2008. Objections to Bayesian statistics. Bayesian Analysis, 3(3), pp. 445-449.
  3. Diaconis, P., 1977. Finite forms of de Finetti's theorem on exchangeability. Synthese, 36(2), pp. 271-281.
  4. Gelman, A., Meng, X. L. and Stern, H., 1996. Posterior predictive assessment of model fitness via realized discrepancies. Statistica sinica, pp. 733-760.
  5. Dunson, D. B., 2018. Statistics in the big data era: Failures of the machine. Statistics & Probability Letters, 136, pp. 4-9.



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




