---
title: "STA 360/602L: Module 2.1"
subtitle: "Conjugacy; beta-Bernoulli and beta-binomial models"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


---
## Outline

- Conjugacy

- Kernels

- Bernoulli data

- Binomial data


---
## Bayesian inference

- Once again, given **data** $y$ and an **unknown population parameter** $\theta$, estimate $\theta$.

--

- As a Bayesian, you update some prior information for $\theta$ with information in the data $y$, to obtain the posterior density $p(\theta | y)$.

--

- Personally, I prefer being able to obtain posterior densities that describe my parameter, instead of estimated summaries (usually measures of central tendency) along with confidence intervals.

--

- Bayes' theorem - reminder: 
.block[
.small[
$$p(\theta | y) = \frac{p(\theta)p(y|\theta)}{\int_{\Theta}p(\tilde{\theta})p(y| \tilde{\theta}) \textrm{d}\tilde{\theta}} = \frac{p(\theta)p(y|\theta)}{p(y)}$$
]
]


---
## Comments on the posterior density

- The posterior density is more concentrated than the prior & quantifies learning about $\theta$.

--

- In fact, this is the optimal way to learn from data - see discussion in Hoff chapter 1.

--

- As more & more data become available, posterior density will converge to a normal (Gaussian) density centered on the MLE (Bayes central limit theorem).

--

- In finite samples for limited data, the posterior can be highly skewed & noticeably non-Gaussian.



---
## Conjugacy

- Starting with an arbitrary prior density $p(\theta)$ & sampling density $p(y|\theta)$ we may encounter problems in calculating the
posterior density $p(\theta | y)$.

--

- In particular, you may notice the denominator in the Bayes' rule: 
.block[
.small[
$$p(y) = \int_{\Theta}p(\theta)p(y| \theta) \textrm{d}\theta.$$
]
]
  This integral may not be analytically tractable!

--

- When the prior is .hlight[conjugate] however, the marginal likelihood can be calculated analytically.

--

- .hlight[Conjugacy] $\Rightarrow$ the posterior density (or mass) function has the same form as the prior density (or mass) function.

--

- Conjugate priors make calculations easy but may not represent our prior information well.



---
## Kernels

- In Bayesian statistics, the .hlight[kernel] of a pdf or pmf omits any multipliers that do not depend on the random variable or parameter we care about.

--

- For many distributions, the kernel is in a simple form but the normalizing constant complicates calculations.

--

- If one recognizes the kernel as that matching a known distribution, then the normalizing factor can be reinstated. This is a very MAJOR TRICK we will use to calculate posterior distributions.

--

- For example, the normal density is given by
.block[
.small[
$$p(y|\mu,\sigma^2) = \dfrac{1}{\sqrt{2\pi\sigma^2}}e^{-\dfrac{(y-\mu)^2}{2\sigma^2}}$$
]
]

--

  but the kernel is just
.block[
.small[
$$p(y|\mu,\sigma^2) \propto e^{-\dfrac{(y-\mu)^2}{2\sigma^2}}.$$
]
]


---
## Bernoulli data

- Back to our example: suppose $\theta \in (0,1)$ is the population proportion of individuals with diabetes in the US.

--

- Suppose we take a sample of $n$ individuals and record whether or not they have diabetes (as binary: 0,1). 

--

- Then we can use the Bernoulli distribution as the sampling distribution. 

--

- Also, we already established that we can use a beta prior for $\theta$.



---
## Bernoulli data

- Generally, it turns out that if 
  + $p(y_i| \theta): y_i \overset{iid}{\sim} \textrm{Bernoulli}(\theta)$ for $i = 1,\ldots,n$, and
  + $\pi(\theta): \theta \sim \textrm{Beta}(a,b)$,
  
  then the posterior distribution is also a beta distribution.
  
--

- <div class="question">
Can we derive the posterior distribution and its parameters? Let's do some work on the board!
</div>

--

- Updating a beta prior with a Bernoulli likelihood leads to a beta posterior - we have conjugacy!

--

- Let $y = (y_1,\ldots,y_n)$. Specifically, we have.
.block[
.small[
$$p(\theta | y) = \textrm{Beta}\left(a+\sum_{i=1}^n y_i,b+n-\sum_{i=1}^n y_i\right).$$
]
]

--

- This is the .hlight[beta-Bernoulli model]. More generally, this is actually the .hlight[beta-binomial model].



---
## Beta-binomial in more detail

- Suppose the sampling density of the data is
.block[
.small[
$$p(y|\theta) = {n \choose y} \theta^y(1-\theta)^{n-y}.$$
]
]


--

- Suppose also that we have a $\textrm{Beta}(a,b)$ prior on the probability $\theta$.

--

- Then the posterior density then has the beta form
.block[
.small[
$$\pi(\theta | y) = \textrm{Beta}(a+y,b+n-y).$$
]
]

--

- The posterior has expectation
.block[
.small[
$$\mathbb{E}(\theta | y) = \dfrac{a+y}{a+b+n} = \dfrac{a+b}{a+b+n} \times \textrm{prior mean} + \dfrac{n}{a+b+n} \times \textrm{sample mean}.$$
]
]

--

- For this specification, **sometimes $a$ and $b$ are interpreted as "prior data" with a interpreted as the prior number of 1's, $b$ as the prior number of 0's, and $a + b$ as the prior sample size.**

--

- As we get more and more data, the majority of our information about $\theta$ comes from the data as opposed to the prior.



---
## Binomial data

- For example, suppose you want to find the Bayesian estimate of the probability $\theta$ that a coin comes up heads.

--

- Before you see the data, you express your uncertainty about $\theta$ through the prior $p(\theta) = \textrm{Beta}(2,2)$

--

- Now suppose you observe 10 tosses, of which only 1 was heads.

--

- Then, the posterior density $p(\theta \,|\, y)$ is $\mbox{Beta}(3, 11)$.



---
## Binomial data

- Recall that the mean of $\mbox{Beta}(a,b)$ is $\frac{a}{a+b}$.

--

- So, before you saw the data, you thought the mean for $\theta$ was $\frac{2}{2+2} = 0.50$. 

--

- However, after seeing the data, you believe it is $\frac{3}{3+11} = 0.21$.

--

- The variance of $\mbox{Beta}(a,b)$ is $\frac{ab}{(a+b)^2(a+b+1)}$.

--

- So before you saw data, your uncertainty about $\theta$, in terms of the standard deviation, was $\sqrt{\frac{4}{4^2 \times 5}} = 0.22$. 

--

- However, after seeing 1 Heads in 10 tosses, your standard deviation gets updated to $\sqrt{\frac{33}{14^2 \times 15}} = 0.11$.

--

- Clearly, as the number of tosses goes to infinity, your uncertainty goes to zero.





