---
title: "STA 360/602L: Module 2.4"
subtitle: "Truncated priors and the inverse cdf method"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Priors with restricted support

- As we have seen, when dealing with rare events, we might expect the true proportion to be very small. 

--

- In that case, we might want to try a restricted prior, e.g. $\text{Unif}(0,0.1)$.

--

- Even when we don't have rare events, we might still desire some restriction if we are certain the true proportion lies within $(a,b)$ with $0 < a < b < 1$.

--

- It is thus often really useful to explore "truncation" on priors.

--

- Let $\theta =$ probability of a randomly-selected student making an $A$ in this course.

--

- You may want to rule out very low & very high values -- perhaps $\theta \in [0.35, 0.6]$ with probability one.

--

- How to choose a prior restricted to this interval?



---
## Uniform priors

- One possibility is to just choose a uniform prior.

--

- When the parameter $\theta$ is a probability, the typical uniform prior would correspond to $\text{Beta}(1,1)$.

--

- This is uniform on the entire $(0,1)$ interval.

--

- However, we can just as easily choose a uniform prior on a narrower interval $\text{Unif}(a,b)$ with $0 < a < b < 1$.

--

- Perhaps not flexible enough.

--

- Would be nice if we could pick a flexible beta density and then truncate it to $(a,b)$.



---
## Truncated random variables

- Suppose we have some arbitrary random variable $\theta \sim f(\theta)$ with support $\Theta$.

--

- For example, $\theta \sim \textrm{Beta}(c,d)$ has support on $(0,1)$.

--

- Then, we can modify the density $f(\theta)$ to have support on a sub-interval $[a,b] \in \Theta$.

--

- The density $f(\theta)$ **truncated** to $[a,b]$ is
.block[
.small[
$$f_{[a,b]}(\theta) = \dfrac{f(\theta)\mathbb{1}[\theta \in [a,b]]}{\int^b_a f(\theta^\star)\textrm{d}\theta^\star},$$
]
]

  with $\mathbb{1}[A]$ being the indicator function that returns 1 if A is true & 0 otherwise.



---
## Truncated beta density

- Suppose to characterize the prior probability of earning an A, you poll a sample of students from a former STA 602
course and find that 10 earned an A and 10 earned a B or lower. 

--

- Therefore, you go with a $\text{Beta}(10,10)$ prior truncated to $[0.35, 0.6]$. 

--

- In R we can calculate the truncated beta density at p via
  ```{r eval=F}
p <- seq(0,1,length=1000)
f1 <- dbeta(p,10,10)
f2 <- dbeta(p,10,10)*as.numeric(p>0.35 & p<0.6)/(pbeta(0.6,10,10) - pbeta(0.3,10,10))
f3 <- dunif(p,0.35,.6)
plot(p,f2,type='l',col='green4',xlim=c(0,1),ylab='Density', xlab=expression(theta),
      ylim=c(0,6))
lines(p,f1,type='l',col='blue')
lines(p,f3,type='l',col='red4')
labels <- c("beta(10,10)", "truncated beta","unif(0.35,.6)")
legend("topright", inset=.05, labels, lwd=2, lty=c(1,1,1), col=c('blue4','green4','red4'))
```



---
## Truncated beta density

What would that look like?
```{r fig.height=5, echo=F}
p <- seq(0,1,length=1000)
f1 <- dbeta(p,10,10)
f2 <- dbeta(p,10,10)*as.numeric(p>0.35 & p<0.6)/ (pbeta(0.6,10,10) - pbeta(0.3,10,10) )
f3 <- dunif(p,0.35,.6)
plot(p,f2,type='l',col='green4',xlim=c(0,1),ylab='Density', xlab=expression(theta),ylim=c(0,6))
lines(p,f1,type='l',col='blue')
lines(p,f3,type='l',col='red4')
labels <- c("beta(10,10)", "truncated beta","unif(0.35,.6)")
legend("topright", inset=.05, labels, lwd=2, lty=c(1,1,1), col=c('blue4','green4','red4'))
```


---
## Truncated beta density

The truncated density by itself would look like
```{r fig.height=5, echo=F}
plot(p,f2,type='l',col='green4',xlim=c(0,1),ylab='Density', xlab=expression(theta),ylim=c(0,6),
     main="Truncated beta")
```



---
## The inverse cdf method

- How to sample truncated random variables?

--

- First start with the pdf for an untruncated distribution such as $\theta \sim \textrm{Beta}(c,d)$.

--

- Suppose we then want to sample $\theta \sim \textrm{Beta}_{[a,b]}(c,d)$. How can we do that? One popular method is the .hlight[inverse-cdf method].

--

- The inverse cdf is useful for generating random variables in general, especially for generating truncated random variables.

--

- Suppose we have $\theta \sim f(\theta)$, for some arbitrary continuous density $f$.

--

- According to probability integral transform, for any continuous random variable $\theta$, the random variable $U = F(\theta)$ has a $\text{Unif}(0,1)$ distribution. Note that $F$ is the cdf.

--

- Thus, to use the inverse-cdf method to sample $\theta \sim f$, first sample $u \sim \textrm{Unif}(0,1)$, then set $\theta = F^{-1}(u)$.



---
## The inverse cdf method

- As an example, suppose we want to sample $\theta \sim \textrm{Beta}(c,d)$ through the inverse cdf method. 

--

- Very easy. Just do the following in R.
  ```{r eval=F}
u <- runif (1, 0, 1)
theta <- qbeta(u,c,d)
```

--

- That is, first sample from a uniform distribution.

--

- Then, transform it using the inverse cdf of the $\textrm{Beta}(c,d)$ distribution.

--

- Viola!



---
## The inverse cdf method

- Back to the original problem: how to sample $\theta \sim \textrm{Beta}_{[a,b]}(c,d)$?

--

- If we had the inverse cdf of $\textrm{Beta}(c,d)$  truncated to $[a, b]$, then we could use the inverse cdf method. Easy enough! Let's find that inverse cdf.

--

- Let $f$, $F$ and $F^{-1}$ denote the pdf, cdf and inverse-cdf without truncation and let $A=[a,b]$.

--

- Recall that the density $f(\theta)$ **truncated** to $[a,b]$ is
.block[
.small[
$$f_{A}(\theta) = f_{[a,b]}(\theta) = \dfrac{f(\theta)\mathbb{1}[\theta \in [a,b]]}{\int^b_a f(\theta^\star)\textrm{d}\theta^\star} = \dfrac{f(\theta)\mathbb{1}[\theta \in [a,b]]}{F(b) - F(a)}.$$
]
]

--

- Therefore, the truncated cdf
.block[
.small[
$$F_{A}(z) = \Pr[\theta \leq z] = \dfrac{F(z) - F(a)}{F(b) - F(a)}.$$
]
]

--

- Not enough though. We need the truncated inverse cdf.



---
## The inverse cdf method

- To find the inverse cdf $F^{-1}_A(u)$, let $F_{A}(z) = u$. That is, set
.block[
.small[
$$u = F_{A}(z) = \dfrac{F(z) - F(a)}{F(b) - F(a)}$$
]
]

  and solve for $z$ as a function of $u$.
  
--

- Re-expressing as a function of $F(z)$,
.block[
.small[
$$F(z) = \{F(b) - F(a)\}u + F(a).$$
]
]

--

- Applying the untruncated inverse cdf $F^{-1}$  to both sides, we have
.block[
.small[
$$z = F^{-1}[\{F(b) - F(a)\}u + F(a)] = F^{-1}_A(u).$$
]
]



---
## The inverse cdf method

- We now have all the pieces to use the inverse-cdf method to sample $\theta \sim f_A$, that is, $f$ truncated to A.

--

- First draw a $\textrm{Unif}(0,1)$ random variable
  ```{r eval=F}
u <- runif (1, 0, 1)
```

--

- Next, apply the linear transformation:
.block[
.small[
$$u^\star = \{F(b) - F(a)\}u + F(a).$$
]
]

--

- Finally, plug $u^\star$ into the untruncated cdf $\theta = F^{-1}(u^\star)$.

--

- Note we can equivalently sample $u^\star \sim runif(1,F(a),F(b))$.






