---
title: "STA 360/602L: Module 2.6"
subtitle: "Loss functions and Bayes risk"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```



## Bayes estimate

- As we've seen by now, having posterior distributions instead of one-number summaries is great for capturing uncertainty.

--

- That said, it is still very appealing to have simple summaries, especially when dealing with clients or collaborators from other fields, who desire one.

--

- Can we obtain a single estimate of $\theta$ based on the posterior? Sure!

--

- .hlight[Bayes estimate] is the value $\hat{\theta}$, that minimizes the Bayes risk.


---
## Bayes estimate

- .hlight[Bayes risk] is defined as the expected loss averaged over the posterior distribution.

--

- Put differently, a .hlight[Bayes estimate] $\hat{\theta}$ has the lowest posterior expected loss.

--

- That's fine, but what does expected loss mean?

--

- .hlight[Frequentist risk] also exists but we won't go into that here.



---
## Loss functions

- A .hlight[loss function] $L(\theta,\delta(y))$ is a function of a parameter $\theta$, where $\delta(y)$ is some .hlight[decision] about $\theta$, based on just the data $y$.

--

- For example, $\delta(y) = \bar{y}$ can be the decision to use the sample mean to estimate $\theta$, the true population mean.

--

- $L(\theta,\delta(y))$ determines the penalty for making the decision $\delta(y)$, if $\theta$ is the true parameter; $L(\theta,\delta(y))$ characterizes the price paid for errors.



---
## Loss functions

- A common choice for example, when dealing with point estimation, is the .hlight[squared error loss], which has
.block[
$$L(\theta,\delta(y)) = (\theta - \delta(y))^2.$$
]

--

- Bayes risk is thus
.block[
$$\rho(\theta,\delta) = \mathbb{E}\left[\ L(\theta,\delta(y)) \ | y\right] = \int L(\theta,\delta(y)) \cdot \pi(\theta|y) \ d\theta,$$
]

  and we proceed to find the value $\hat{\theta}$, that is, the decision $\delta(y)$, that minimizes the Bayes risk.




---
## Bayes estimator under squared error loss

- Turns out that, under squared error loss, the decision $\delta(y)$ that minimizes the posterior risk is the posterior mean.

--

- Proof: Let $L(\theta,\delta(y)) = (\theta - \delta(y))^2$. Then,
.block[
$$
\begin{aligned}
\rho(\theta,\delta) & = \int L(\theta,\delta(y)) \cdot \pi(\theta|y) \ d\theta.\\
& = \int (\theta - \delta(y))^2 \cdot \pi(\theta|y) \ d\theta.\\
\end{aligned}
$$
]

--

- Expand, then take the partial derivative of $\rho(\theta,\delta)$ with respect to $\delta(y)$.

--

- <div class="question">
To be continued on the board!
</div>


---
## Bayes estimator under squared error loss

- .block[
$$
\begin{aligned}
\rho(\theta,\delta) & \int (\theta - \delta(y))^2 \cdot \pi(\theta|y) \ d\theta.\\
\end{aligned}
$$
]

- Easy to see then that $\delta(y) = \mathbb{E}[\theta | x]$ is the minimizer.

--

- Well that's great! The posterior mean is often very easy to calculate in most cases. 

--

- In the beta-binomial case for example, the Bayes estimate under squared error loss is just
.block[
$$\hat{\theta} = \frac{a+y}{a+b+n},$$
]

  the posterior mean.



---
## What about other loss functions?

- Clearly, squared error is only one possible loss function. An alternative is .hlight[absolute loss], which has
.block[
$$L(\theta,\delta(y)) = |\theta - \delta(y)|.$$
]

--

- Absolute loss places less of a penalty on large deviations & the resulting Bayes estimate is **posterior median**.

--

- Median is actually relatively easy to estimate.


---
## What about other loss functions?

- Recall that for a continuous random variable $Y$ with cdf $F$, the median of the distribution is the value $z$, which satisfies
.block[
$$F(z) = \Pr(Y\leq z) = \dfrac{1}{2}= \Pr(Y\geq z) = 1-F(z).$$
]

--

- As long as we know how to evaluate the CDF of the distribution we have, we can solve for $z$. 

--

- Think R!



---
## What about other loss functions?

- For the beta-binomial model, the CDF of the beta posterior can be written as
.block[
$$F(z) = \Pr(\theta\leq z | y) = \int^z_0 \textrm{Beta}(\theta| a+y, b+n-y) d\theta.$$
]

--

- Then, if $\hat{\theta}$ is the median, we have that $F(\hat{\theta}) = 0.5$.

--

- To solve for $\hat{\theta}$, apply the inverse CDF $\hat{\theta} = F^{-1}(0.5)$.

--

- In R, that's simply
    ```{r eval=F}
qbeta(0.5,a+y,b+n-y)
```

--

- For other popular distributions, switch out the beta.



---
## Loss functions and decisions

- Loss functions are not specific to estimation problems but are a critical part of decision making.

--

- For example, suppose you are deciding how much money to bet ($A) on Duke in the next UNC-Duke men's basketball game.

--

- Suppose, if Duke
  + loses (y = 0), you lose the amount you bet ($A)
  + wins (y = 1), you gain B per $1 bet
  
--

- <div class="question">
What is a good sampling distribution for y here?
</div>
  
--

- Then, the loss function can be characterized as
.block[
.small[
$$L(A,y) = A(1-y) - y(BA),$$
]
]
  
  with your action being the amount bet A.
  
--

- <div class="question">
When will your bet be "rational"?
</div>



---
## How much to bet on Duke?

- $y$ is an unknown state, but we can think of it as a new prediction $y_{n+1}$ given that we have data from win-loss records $(y_{1:n})$ that can be converted into a Bayesian posterior,
.block[
.small[
$$\theta \sim \textrm{beta}(a_n,b_n),$$
]
]

--
  with this posterior concentrated slightly to the left of 0.5, if we only use data on UNC-Duke games (UNC men lead Duke 139-112 all time).
  
--

- Actually, it might make more sense to focus on more recent head-to-head data and not the all time record.

--

- In fact, we might want to build a model that predicts the outcome of the game using historical data & predictors (current team rankings, injuries, etc).

--

- However, to keep it simple for this illustration, go with the posterior above.



---
## How much to bet on Duke?

- The Bayes risk for action A is then the expectation of the loss function,
.block[
.small[
$$\rho(A) = \mathbb{E}\left[\ L(A,y) | \ y_{1:n}\right].$$
]
]

--

- To calculate this as a function of $A$ and find the optimal $A$, we need to marginalize over the **posterior predictive distribution** for $y$.

--

- <div class="question">
Why are we using the posterior predictive distribution here instead of the posterior distribution?
</div>

--

- As an aside, recall from Module 2.3 that
.block[
.small[
$$p(y_{n+1}|y_{1:n}) = \dfrac{a_n^{y_{n+1}} b_n^{1-y_{n+1}}}{a_n + b_n}; \ \ \ y_{n+1}=0,1.$$
]
]

--

- Specifically, that the posterior predictive distribution here is $\textrm{Bernoulli}(\hat{\theta})$, with
.block[
.small[
$$\hat{\theta} = \dfrac{a_n}{a_n + b_n}$$
]
]

--

- By the way, what do $a_n$ and $b_n$ represent?



---
## How much to bet on Duke?

- With the loss function $L(A,y) = A(1-y) - y(BA)$, and using the notation $y_{n+1}$ instead of $y$ (to make it obvious the game has not been played), the Bayes risk (expected loss) for bet $A$ is
.block[
$$
\begin{split}
\rho(A) & = \mathbb{E}\left[\ L(A,y_{n+1}) \ | y_{1:n}\right]\\
\\
& = \mathbb{E}\left[A(1-y_{n+1}) - y_{n+1}(BA) \ | y_{1:n}\right]\\
\\
& = A \ \mathbb{E}\left[1-y_{n+1} | \ y_{1:n}\right]  - (BA) \ \mathbb{E}\left[y_{n+1} | y_{1:n}\right]\\
\\
& = A \ \left(1 -  \mathbb{E}\left[y_{n+1} | y_{1:n}\right] \right)  - (BA) \ \mathbb{E}\left[y_{n+1} | y_{1:n}\right]\\
\\
& = A \ \left(1 -\mathbb{E}\left[y_{n+1} | y_{1:n}\right] \ (1+B) \right).
\end{split}
$$
]


---
## How much to bet on Duke?

- Hence, your bet is rational as long as
.block[
$$
\begin{split}
\mathbb{E}\left[y_{n+1} | \ y_{1:n}\right](1+B) > 1\\
\\
\dfrac{a_n (1+B)}{a_n + b_n} > 1.
\end{split}
$$
]

--

- Clearly, there is no limit to the amount you should bet if this condition is satisfied (the loss function is clearly too simple).

--

- Loss function needs to be carefully chosen to lead to a good decision - finite resources, diminishing returns, limits on donations, etc.

--

- Want more on loss functions, expected loss/utility, or decision problems in general? Consider taking a course on decision theory (STA623?).







