---
title: "STA 360/602L: Module 3.11"
subtitle: "Discussion session exercise"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## In-class exercise

- Your friend agrees to conduct a poll for you, free of charge (lucky you!). 

--

- You give the following instructions: “Please ask about 25 people whether they are in favor of more gun control, and report back to me the number who are in favor.”

--

- After a few days your friend returns with the poll results: there were $y = 20$ in favor. “

--

- You then ask, "how many people did you ask?”Your friend says, "ummm, I dunno. You didn't ask me to record that. All I know is that it was about 25.”

--

- <div class="question">
What model can we use to do inference here?
</div>

--

- **To be done during the discussion session.**



---
## Exercise

- Do the following:

  1. Using the full conditionals derived during the session, write a Gibbs sampler to sample from the joint posterior of $N$ and $\theta$, using a starting value of $N = 50$ and $\theta = 0.05$. Set burn-in to $2000$ and then proceed to generate $10000$ draws.

  2. Look at the posterior densities for both parameters. Describe the distributions.

  3. Give the quantile-based equal-tailed posterior credible interval for $\theta$, rounded to two decimal places.

  4. What is the probability that exactly $20$ people were polled? What can you takeaway from this?

  5. What is the probability that exactly $25$ people were polled? What can you takeaway from this?




