---
title: "STA 360/602L: Module 3.4"
subtitle: "The normal model: conditional inference for the mean"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Normal model

- Suppose we have independent observations $Y = (y_1,y_2,\ldots,y_n)$, where each $y_i \sim \mathcal{N}(\mu, \sigma^2)$ or $y_i \sim \mathcal{N}(\mu, \tau^{-1})$, with unknown parameters $\mu$ and $\sigma^2$ (or $\tau$).

--

- Then, the likelihood is
.block[
.small[
$$
\begin{split}
P(Y| \mu,\sigma^2) & = \prod_{i=1}^n \dfrac{1}{\sqrt{2\pi}} \tau^{\frac{1}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau (y_i-\mu)^2\right\}\\
& \propto \tau^{\frac{n}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau \sum_{i=1}^n (y_i-\mu)^2\right\}\\
& \propto \tau^{\frac{n}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau \sum_{i=1}^n \left[ (y_i-\bar{y}) - (\mu - \bar{y}) \right]^2 \right\}\\
\\
& \propto \tau^{\frac{n}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau \left[ \sum_{i=1}^n (y_i-\bar{y})^2 - \sum_{i=1}^n(\mu - \bar{y})^2 \right] \right\}\\
& \propto \tau^{\frac{n}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau \left[ \sum_{i=1}^n (y_i-\bar{y})^2 - n(\mu - \bar{y})^2 \right] \right\}\\
& \propto \tau^{\frac{n}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau s^2(n-1) \right\} \ \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu - \bar{y})^2 \right\}.\\
\end{split}
$$
]
]



---
## Likelihood for normal model

- Likelihood:
.block[
.large[
$$P(Y| \mu,\sigma^2) \propto \tau^{\frac{n}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau s^2(n-1) \right\} \ \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu - \bar{y})^2 \right\},$$
]
]

  where
  + $\bar{y} =\sum_{i=1}^n y_i$ is the sample mean; and
  + $s^2 = \sum_{i=1}^n (y_i-\bar{y})^2/(n-1)$ is the sample variance.

- Sufficient statistics:
  + Sample mean $\bar{y}$; and
  + Sample sum of squares $SS = s^2(n-1) = \sum_{i=1}^n (y_i-\bar{y})^2$.

--

- MLEs:
  + $\hat{\mu} = \bar{y}$.
  + $\hat{\tau} = n/SS$, and $\hat{\sigma}^2 = SS/n$.



---
## Inference for mean, conditional on variance

- We can break down inference problem for this two-parameter model into two one-parameter problems.

--

- First start by developing inference on $\mu$ when $\sigma^2$ is known. Turns out we can use a conjugate prior for $\pi(\mu|\sigma^2)$. We will get to unknown $\sigma^2$ in the next module.

--

- For $\sigma^2$ known, the normal likelihood further simplifies to
.block[
.small[
$$\propto \ \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu - \bar{y})^2 \right\},$$
]
]

  leaving out everything else that does not depend on $\mu$.
  
--

- For $\pi(\mu|\sigma^2)$, we consider $\mathcal{N}(\mu_0, \sigma_0^2)$, i.e., $\mathcal{N}(\mu_0, \tau_0^{-1})$, where $\tau_0^{-1} = \sigma_0^2$.

--

- Let's derive the posterior $\pi(\mu|Y,\sigma^2)$.


---
## Inference for mean, conditional on variance

- First, the prior $\pi(\mu|\sigma^2) = \mathcal{N}(\mu_0, \tau_0^{-1})$ can be written as
.block[
.small[
$$
\begin{split}
\Rightarrow \pi(\mu|\sigma^2) \ & = \ \dfrac{1}{\sqrt{2\pi}} \tau_0^{\frac{1}{2}} \cdot \textrm{exp}\left\{-\frac{1}{2} \tau_0 (\mu-\mu_0)^2) \right\} \\
\\
&  \propto \ \textrm{exp}\left\{-\frac{1}{2} \tau_0 (\mu^2 - 2\mu\mu_0 + \mu_0^2) \right\} \\
\\
&  \propto \ \textrm{exp}\left\{-\frac{1}{2} \tau_0 (\mu^2 - 2\mu\mu_0) \right\}.\\
\end{split}
$$
]
]

--

- **When the normal density is written in this form, note the following details in the exponent.**

  + First, we must have $\mu^2 - 2\mu$, and whatever term we see multiplying $2\mu$ must be the mean, in this case, $\mu_0$.
  
  + Second, the precision $\tau_0$ is outside the parenthensis.



---
## Inference for mean, conditional on variance

- Now to the posterior:
.block[
.small[
$$\pi(\mu|Y,\sigma^2) \ \propto \ \pi(\mu|\sigma^2) P(Y| \mu,\sigma^2) \ \propto \ \textrm{exp}\left\{-\frac{1}{2} \tau_0 (\mu - \mu_0)^2 \right\}\  \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu - \bar{y})^2 \right\}$$
]
]

--

- Expanding out squared terms
.block[
.small[
$$\Rightarrow \pi(\mu|Y,\sigma^2) \ \propto \ \textrm{exp}\left\{-\frac{1}{2} \tau_0 (\mu^2 - 2\mu\mu_0 + \mu_0^2) \right\}\  \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu^2 - 2\mu\bar{y} + \bar{y}^2) \right\}$$
]
]

--

- Ignoring terms not containing $\mu$
.block[
.small[
$$
\begin{split}
\Rightarrow \pi(\mu|Y,\sigma^2) \ & \propto \ \textrm{exp}\left\{-\frac{1}{2} \tau_0 (\mu^2 - 2\mu\mu_0) \right\}\  \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu^2 - 2\mu\bar{y}) \right\}\\
\\
& = \ \textrm{exp}\left\{-\frac{1}{2} \left[\tau_0 (\mu^2 - 2\mu\mu_0)  + \tau n(\mu^2 - 2\mu\bar{y}) \right] \right\}\\
\\
& = \ \textrm{exp}\left\{-\frac{1}{2} \left[ \mu^2(\tau n + \tau_0) - 2\mu(\tau n\bar{y} + \tau_0\mu_0)   \right] \right\}.\\
\end{split}
$$
]
]




---
## Inference for mean, conditional on variance

- This sort of looks like a normal kernel but we need to do a bit more work to get there.

--

- Particularly, we need to have it be of the form $b(\mu^2 - 2\mu a)$, so that we have $a$ as the mean and $b$ as the precision.

--

- We have
.block[
.small[
$$
\begin{split}
\pi(\mu|Y,\sigma^2) \ & \propto \ \textrm{exp}\left\{-\frac{1}{2} \left[ \mu^2(\tau n + \tau_0) - 2\mu(\tau n\bar{y} + \tau_0\mu_0)   \right] \right\}\\
\\
& = \ \textrm{exp}\left\{-\frac{1}{2} \cdot (\tau n + \tau_0) \left[ \mu^2 - 2\mu \left( \frac{\tau n\bar{y} + \tau_0\mu_0}{\tau n + \tau_0} \right)   \right] \right\}.\\
\end{split}
$$
]
]

  which now looks like the kernel of a normal distribution.



---
## Posterior with precision terms

- Again, the posterior is
.block[
$$
\begin{split}
\pi(\mu|Y,\sigma^2) \ & \propto \ \textrm{exp}\left\{-\frac{1}{2} \cdot (\tau n + \tau_0) \left[ \mu^2 - 2\mu \left( \frac{\tau n\bar{y} + \tau_0\mu_0}{\tau n + \tau_0} \right)   \right] \right\}.\\
\end{split}
$$
]

--

- So, in terms of precision, we have
.block[
$$\mu|Y,\sigma^2 \sim \mathcal{N}(\mu_n, \tau_n^{-1})$$
]

  where
.block[
$$\mu_n = \dfrac{\tau n\bar{y} + \tau_0\mu_0}{\tau n + \tau_0}$$
]

  and
.block[
$$\tau_n = \tau n + \tau_0.$$
]



---
## Posterior with precision terms

- As mentioned before, Bayesians often prefer to talk about precision instead of variance.

--

- We have
  + $\tau$ as the sampling precision (how close the $y_i$'s are to $\mu$).
  
  + $\tau_0$ as the prior precision (our prior belief about the uncertainty about $\mu$ around our prior guess $\mu_0$).
  
  + $\tau_n$ as the posterior precision

--

- From the posterior, we can see that, _the posterior precision equals the prior precision plus the data precision_.

--

- That is, once again, the posterior information is a combination of the prior information and the information from the data.



---
## Posterior with precision terms: combining information

- Posterior mean is weighted sum of prior information plus data information:
.block[
$$
\begin{split}
\mu_n & = \dfrac{n\tau\bar{y} + \tau_0\mu_0}{\tau n + \tau_0}\\
& = \dfrac{\tau_0}{\tau_0 + \tau n} \mu_0 + \dfrac{n\tau}{\tau_0 + \tau n} \bar{y}
\end{split}
$$
]

--

- Recall that $\sigma^2$ (and thus $\tau$) is known for now. 

--

- If we think of the prior mean as being based on $\kappa_0$ prior observations from a similar population as $y_1,y_2,\ldots,y_n$, then we might set $\sigma_0^2 = \frac{\sigma^2}{\kappa_0}$, which implies $\tau_0 = \kappa_0 \tau$, and then the posterior mean is given by
.block[
$$
\begin{split}
\mu_n & = \dfrac{\kappa_0}{\kappa_0 + n} \mu_0 + \dfrac{n}{\kappa_0 + n} \bar{y}.
\end{split}
$$
]


---
## Posterior with variance terms

- In terms of variances, we have
.block[
$$\mu|Y,\sigma^2 \sim \mathcal{N}(\mu_n, \sigma_n^2)$$
]

  where
.block[
$$\mu_n = \dfrac{ \dfrac{n}{\sigma^2}\bar{y} + \dfrac{1}{\sigma^2_0} \mu_0}{\dfrac{n}{\sigma^2} + \dfrac{1}{\sigma^2_0}}$$
]

  and
.block[
$$\sigma^2_n = \dfrac{1}{\dfrac{n}{\sigma^2} + \dfrac{1}{\sigma^2_0}}.$$
]

--

- It is still easy to see that we can re-express the posterior information as a sum of the prior information and the information from the data.








---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




