---
title: "STA 360/602L: Module 5.1"
subtitle: "Hierarchical normal models with constant variance: two groups"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(lattice)
library(MCMCpack)
library(hdrcde)
library(coda)
library(rethinking)
```



## Motivation

- Sometimes, we may have a natural grouping in our data, for example 
  + students within schools,
  + patients within hospitals,
  + voters within counties or states,
  + biology data, where animals are followed within natural populations organized geographically and, in some cases, socially.

--

- For such grouped data, we may want to do inference across all the groups, for example, comparison of the group means.

--

- Ideally, we should do so in a way that takes advantage of the relationship between observations in the same group, but we should also look to borrow information across groups when possible.

--

- .hlight[Hierarchical modeling] provides a principled way to do so. 



---
## Bayes estimators and bias

- Recall the normal model:
.block[
.small[
$$y_{i} | \mu, \sigma^2 \overset{iid}{\sim} \mathcal{N} \left( \mu, \sigma^2\right).$$
]
]

--

- The MLE for the population mean $\mu$ is just the sample mean $\bar{y}$.

--

- $\bar{y}$ is unbiased for $\mu$. That is, for any data $y_i \overset{iid}{\sim} \mathcal{N} \left( \mu, \sigma^2\right)$, $\mathbb{E}[\bar{y}] = \mu$.

--

- However, recall that in the conjugate normal model with known variance for example, the posterior expectation is a **weighted average** of the prior mean and the sample mean.

--

- That is, the posterior mean is actually biased.



---
## Shrinkage

- Usually through the weighting of the sample data and prior, Bayes procedures have the tendency to pull the estimate of $\mu$ toward the prior
mean.

--

- Of course, the magnitude of the pull depends on the sample size.

--

- This "pulling" phenomenon is referred to as .hlight[shrinkage].

--

- <div class="question">
Why would we ever want to do this? Why not just stick with the MLE?
</div>

--

- Well, in part, because shrinkage estimators are often "more accurate" in prediction problems -- i.e. they tend to do a
better job of predicting a future outcome or of recovering the actual parameter values. Remember variance-bias trade off!

--

- The fact that a biased estimator would do a better job in many prediction problems can be proven rigorously, and is referred to as .hlight[Stein's paradox].



---
## Modern relevance

- Stein's result implies, in particular, that the sample mean is an *inadmissible* estimator of the mean of a multivariate normal distribution in more than two dimensions -- i.e. there are other estimators that will come closer to the true value in expectation.

--

- In fact, these are Bayes point estimators (the posterior expectation of the parameter $\mu$).

--

- Most of what we do now in high-dimensional statistics is develop biased estimators that perform better than unbiased ones.

--

- Examples: lasso regression, ridge regression, various kinds of hierarchical Bayesian models, etc.

--

- So, here we will get a very basic introduction to .hlight[Bayesian hierarchical models], which provide a formal and coherent framework for constructing shrinkage estimators.



---
## Why hierarchical models?

- **Bayesian hierarchical models** is a sort of catch-all phrase for a large class of models that have several levels of conditional distributions making up the prior.

--

- Like simpler one-level priors, they also accomplish shrinkage. However, they are much more flexible.

--

- Why use them? Several reasons:

  + We may want to exploit more complex dependence structures.

--
  + We may have many parameters relative to the amount of data that we have, and want to borrow information in estimating them.
  
--
  + We may want to shrink toward something other than a simple prior mean/hyper-parameter.



---
## Comparing two groups

- Suppose we want to do inference on mean body mass index (BMI) for two groups (male or female).

--

- BMI is known to often follow a normal distribution, so let's assume the same here.

--

- We should expect some relationship between the mean BMI for the two groups. 

--

- We may also think the shape of the two distributions would be relatively the same (at least as a simplifying assumption for now). 

--

- Thus, a reasonable model might be
.block[
.small[
$$
\begin{split}
y_{i,male}  & \overset{iid}{\sim} \mathcal{N} \left(\theta_m, \sigma^2\right); \ \ i = 1, \ldots, n_m;\\
y_{i,female}  & \overset{iid}{\sim} \mathcal{N} \left(\theta_f, \sigma^2\right); \ \ i = 1, \ldots, n_f.\\
\end{split}
$$
]
]

  but with some relationship between $\theta_m$ and $\theta_f$.


---
## Bayesian inference

- One parameterization  that can reflect some relationship between $\theta_m$ and $\theta_f$ is
.block[
.small[
$$
\begin{split}
y_{i,male}  & \overset{iid}{\sim} \mathcal{N} \left(\mu + \delta, \sigma^2\right); \ \ i = 1, \ldots, n_m;\\
y_{i,female}  & \overset{iid}{\sim} \mathcal{N} \left(\mu - \delta, \sigma^2\right); \ \ i = 1, \ldots, n_f.\\
\end{split}
$$
]
]

--

  where
    + $\theta_m = \mu + \delta$ and $\theta_f = \mu - \delta$,
    
    + $\mu = \dfrac{\theta_m + \theta_f}{2}$ is the average of the population means, and 
    
    + $2\delta = \theta_m - \theta_f$ is the difference in population means.

---
## Bayesian inference

- Convenient prior: 
  + $\pi(\mu, \delta, \sigma^2) = \pi(\mu) \cdot \pi(\delta) \cdot \pi(\sigma^2)$, where
      - $\pi(\mu) = \mathcal{N}(\mu_0, \gamma_0^2)$,
      
      - $\pi(\delta) = \mathcal{N}(\delta_0, \tau_0^2)$, and
      
      - $\pi(\sigma^2) = \mathcal{IG}(\dfrac{\nu_0}{2}, \dfrac{\nu_0 \sigma_0^2}{2})$.
      
--

- We will set the hyper-parameters as:
  + $\mu_0 = 15, \gamma_0 = 5$,
  + $\delta_0 = 0, \tau_0 = 3$,
  + $\nu_0 = 1, \sigma_0 = 5$.

--

- <div class="question">
Do these values seem reasonable?
</div>



---
## Bayesian inference

- Note that we can rewrite
.block[
.small[
$$
\begin{split}
y_{i,male}  & \overset{iid}{\sim} \mathcal{N} \left(\mu + \delta, \sigma^2\right); \ \ i = 1, \ldots, n_m;\\
y_{i,female}  & \overset{iid}{\sim} \mathcal{N} \left(\mu - \delta, \sigma^2\right); \ \ i = 1, \ldots, n_f\\
\end{split}
$$
]
]

  as
.block[
.small[
$$
\begin{split}
(y_{i,male} - \delta) & \overset{iid}{\sim} \mathcal{N} \left(\mu, \sigma^2\right); \ \ i = 1, \ldots, n_m;\\
(y_{i,female} + \delta)  & \overset{iid}{\sim} \mathcal{N} \left(\mu, \sigma^2\right); \ \ i = 1, \ldots, n_f\\
\end{split}
$$
]
]

--

  or
.block[
.small[
$$
\begin{split}
(y_{i,male} - \mu) & \overset{iid}{\sim} \mathcal{N} \left(\delta, \sigma^2\right); \ \ i = 1, \ldots, n_m;\\
(-1) (y_{i,female} - \mu)  & \overset{iid}{\sim} \mathcal{N} \left(\delta, \sigma^2\right); \ \ i = 1, \ldots, n_f.\\
\end{split}
$$
]
]

  as needed, so we can leverage past results for the full conditionals.



---
## Full conditionals

- For the full conditionals we will derive here, we will take advantage of previous results from the regular univariate normal model.

--

- Recall that if we assume
.block[
.small[
$$y_i \sim \mathcal{N}(\mu, \sigma^2), \ \ i=1, \ldots, n,$$
]
]

--

  and set our priors to be
.block[
.small[
$$
\begin{split}
\pi(\mu) & = \mathcal{N}\left(\mu_0, \gamma_0^2\right).\\
\pi(\sigma^2) & = \mathcal{IG}\left(\dfrac{\nu_0}{2}, \dfrac{\nu_0 \sigma_0^2}{2}\right),\\
\end{split}
$$
]
]

--

  then we have
.block[
.small[
$$
\begin{split}
\pi(\mu, \sigma^2 | Y) & \boldsymbol{\propto} \left\{ \prod_{i=1}^{n} p(y_{i} | \mu, \sigma^2 ) \right\} \cdot \pi(\mu) \cdot \pi(\sigma^2) \\
\end{split}
$$
]
]



---
## Full conditionals

- We have
.block[
.small[
$$
\begin{split}
\pi(\mu | \sigma^2, Y) & = \mathcal{N}\left(\mu_n, \gamma_n^2\right).\\
\end{split}
$$
]
]

  where
.block[
.small[
$$
\begin{split}
\gamma^2_n = \dfrac{1}{ \dfrac{n}{\sigma^2} + \dfrac{1}{\gamma_0^2} }; \ \ \ \ \ \ \ \ \mu_n & = \gamma^2_n \left[ \dfrac{n}{\sigma^2} \bar{y} + \dfrac{1}{\gamma_0^2} \mu_0 \right],
\end{split}
$$
]
]

--

- and
.block[
.small[
$$
\begin{split}
\pi(\sigma^2 | \mu,Y) \boldsymbol{=} \mathcal{IG}\left(\dfrac{\nu_n}{2}, \dfrac{\nu_n\sigma_n^2}{2}\right),
\end{split}
$$
]
]

  where
.block[
.small[
$$
\begin{split}
\nu_n & = \nu_0 + n; \ \ \ \ \ \ \ \sigma_n^2  = \dfrac{1}{\nu_n} \left[ \nu_0 \sigma_0^2 + \sum^n_{i=1} (y_i - \mu)^2 \right].\\
\end{split}
$$
]
]





---
## Full conditionals

- With $\pi(\mu) = \mathcal{N}(\mu_0, \gamma_0^2)$, and
.block[
.small[
$$
\begin{split}
(y_{i,male} - \delta) & \overset{iid}{\sim} \mathcal{N} \left(\mu, \sigma^2\right); \ \ i = 1, \ldots, n_m;\\
(y_{i,female} + \delta)  & \overset{iid}{\sim} \mathcal{N} \left(\mu, \sigma^2\right); \ \ i = 1, \ldots, n_f,\\
\end{split}
$$
]
]

  we have
.block[
.small[
$$
\begin{split}
\mu | Y, \delta, \sigma^2 & \sim \mathcal{N}(\mu_n, \gamma_n^2), \ \ \ \text{where}\\
\\
\gamma_n^2 & = \dfrac{1}{\dfrac{1}{\gamma_0^2} + \dfrac{n_m + n_f}{\sigma^2} }\\
\\
\mu_n & = \gamma_n^2 \left[\dfrac{\mu_0}{\gamma_0^2} + \dfrac{ \sum\limits_{i=1}^{n_m} (y_{i,male} - \delta) + \sum\limits_{i=1}^{n_f} (y_{i,female} + \delta) }{\sigma^2} \right].\\
\end{split}
$$
]
]


---
## Full conditionals

- With $\pi(\delta) = \mathcal{N}(\delta_0, \tau_0^2)$, and
.block[
.small[
$$
\begin{split}
(y_{i,male} - \mu) & \overset{iid}{\sim} \mathcal{N} \left(\delta, \sigma^2\right); \ \ i = 1, \ldots, n_m;\\
(-1) (y_{i,female} - \mu)  & \overset{iid}{\sim} \mathcal{N} \left(\delta, \sigma^2\right); \ \ i = 1, \ldots, n_f,\\
\end{split}
$$
]
]

  we have
.block[
.small[
$$
\begin{split}
\delta | Y, \mu, \sigma^2 & \sim \mathcal{N}(\delta_n, \tau_n^2), \ \ \ \text{where}\\
\\
\tau_n^2 & = \dfrac{1}{\dfrac{1}{\tau_0^2} + \dfrac{n_m + n_f}{\sigma^2} }\\
\\
\delta_n & = \tau_n^2 \left[\dfrac{\delta_0}{\tau_0^2} + \dfrac{\sum\limits_{i=1}^{n_m} (y_{i,male} - \mu) + (-1) \sum\limits_{i=1}^{n_f} (y_{i,female} - \mu) }{\sigma^2} \right].\\
\end{split}
$$
]
]

---
## Full conditionals

- With $\pi(\sigma^2) = \mathcal{IG}(\dfrac{\nu_0}{2}, \dfrac{\nu_0 \sigma_0^2}{2})$, and
.block[
.small[
$$
\begin{split}
y_{i,male}  & \overset{iid}{\sim} \mathcal{N} \left(\mu + \delta, \sigma^2\right); \ \ i = 1, \ldots, n_m;\\
y_{i,female}  & \overset{iid}{\sim} \mathcal{N} \left(\mu - \delta, \sigma^2\right); \ \ i = 1, \ldots, n_f\\
\end{split}
$$
]
]

  we have
.block[
.small[
$$
\begin{split}
\sigma^2 | Y, \mu, \delta & \sim \mathcal{IG}(\dfrac{\nu_n}{2}, \dfrac{\nu_n \sigma_n^2}{2}), \ \ \ \text{where}\\
\\
\nu_n & = \nu_0 + n_m + n_f\\
\\
\sigma_n^2 & = \dfrac{1}{\nu_n} \left[\nu_0\sigma_0^2 + \sum\limits_{i=1}^{n_m} (y_{i,male} - [\mu + \delta])^2 + \sum\limits_{i=1}^{n_f} (y_{i,female} - [\mu - \delta])^2 \right].\\
\end{split}
$$
]
]

--

- We will use write a Gibbs sampler for this model and fit the model to real data in the next module.


---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




