---
title: "STA 360/602L: Module 5.2"
subtitle: "Hierarchical normal models with constant variance: two groups (illustration)"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(lattice)
library(MCMCpack)
library(hdrcde)
library(coda)
library(rethinking)
```


class: center, middle

### No pre-recorded video for this module. To be done during discussion session.


---
## Full conditionals recap

.block[
.small[
$$
\begin{split}
\mu | Y, \delta, \sigma^2 & \sim \mathcal{N}(\mu_n, \gamma_n^2), \ \ \ \text{where}\\
\\
\gamma_n^2 & = \dfrac{1}{\dfrac{1}{\gamma_0^2} + \dfrac{n_m + n_f}{\sigma^2} }\\
\\
\mu_n & = \gamma_n^2 \left[\dfrac{\mu_0}{\gamma_0^2} + \dfrac{ \sum\limits_{i=1}^{n_m} (y_{i,male} - \delta) + \sum\limits_{i=1}^{n_f} (y_{i,female} + \delta) }{\sigma^2} \right].\\
\end{split}
$$
]
]


---
## Full conditionals

.block[
.small[
$$
\begin{split}
\delta | Y, \mu, \sigma^2 & \sim \mathcal{N}(\delta_n, \tau_n^2), \ \ \ \text{where}\\
\\
\tau_n^2 & = \dfrac{1}{\dfrac{1}{\tau_0^2} + \dfrac{n_m + n_f}{\sigma^2} }\\
\\
\delta_n & = \tau_n^2 \left[\dfrac{\delta_0}{\tau_0^2} + \dfrac{\sum\limits_{i=1}^{n_m} (y_{i,male} - \mu) + (-1) \sum\limits_{i=1}^{n_f} (y_{i,female} - \mu) }{\sigma^2} \right].\\
\end{split}
$$
]
]

---
## Full conditionals

.block[
.small[
$$
\begin{split}
\sigma^2 | Y, \mu, \delta & \sim \mathcal{IG}(\dfrac{\nu_n}{2}, \dfrac{\nu_n \sigma_n^2}{2}), \ \ \ \text{where}\\
\\
\nu_n & = \nu_0 + n_m + n_f\\
\\
\sigma_n^2 & = \dfrac{1}{\nu_n} \left[\nu_0\sigma_0^2 + \sum\limits_{i=1}^{n_m} (y_{i,male} - [\mu + \delta])^2 + \sum\limits_{i=1}^{n_f} (y_{i,female} - [\mu - \delta])^2 \right].\\
\end{split}
$$
]
]



---
## Application to data

- The data we will use in the $\textsf{R}$ package `rethinking`.
  ```{r}
#install.packages(c("coda","devtools","loo","dagitty"))
#library(devtools)
#devtools::install_github("rmcelreath/rethinking",ref="Experimental")
#library(rethinking)
data(Howell1)

Howell1[1:15,]
```



---
## Application to data

- For now, focus on data for individuals under age 15.
  ```{r}
htm <- Howell1$height/100
bmi <- Howell1$weight/(htm^2)
y_male <- bmi[Howell1$age<15 & Howell1$male==1]
y_female <- bmi[Howell1$age<15 & Howell1$male==0]
n_m <- length(y_male)
n_f <- length(y_female)

n_f
n_m

summary(y_male)
summary(y_female)
```



---
## Application to data

- We will set the hyper-parameters as:
  + $\mu_0 = 15, \gamma_0 = 5$,
  
  + $\delta_0 = 0, \tau_0 = 3$,
  
  + $\nu_0 = 1, \sigma_0 = 5$.

--

- <div class="question">
Do these values seem reasonable?
</div>




---
## Application to data

```{r}
#priors
mu0 <- 15; gamma02 <- 5^2
delta0 <- 0; tau02 <- 3^2
nu0 <- 1; sigma02 <- 5^2

#starting values
mu <- (mean(y_male) + mean(y_female))/2
delta <- (mean(y_male) - mean(y_female))/2
#no need for starting values for sigma_squared, we can sample it first

MU <- DELTA <- SIGMA2 <- NULL
```




---
## Application to data

```{r, cache=T}
#set seed
set.seed(1234)

#set number of iterations and burn-in
n_iter <- 10000; burn_in <- 0.2*n_iter

##Gibbs sampler
for (s in 1:(n_iter+burn_in)) {
#update sigma2
sigma2 <- 1/rgamma(1,(nu0 + n_m + n_f)/2,
                   (nu0*sigma02 + sum((y_male-mu-delta)^2) + sum((y_female-mu+delta)^2))/2)

#update mu
gamma2n <- 1/(1/gamma02 + (n_m + n_f)/sigma2)
mun <- gamma2n*(mu0/gamma02 + sum(y_male-delta)/sigma2 + sum(y_female+delta)/sigma2)
mu <- rnorm(1,mun,sqrt(gamma2n))

#update delta
tau2n <- 1/(1/tau02 + (n_m+n_f)/sigma2)
deltan <- tau2n*(delta0/tau02 + sum(y_male-mu)/sigma2 - sum(y_female-mu)/sigma2)
delta <- rnorm(1,deltan,sqrt(tau2n))

#save parameter values
MU <- c(MU,mu); DELTA <- c(DELTA,delta); SIGMA2 <- c(SIGMA2,sigma2)
}
```



---
## Posterior summaries

```{r fig.height=4}
#library(coda)
MU.mcmc <- mcmc(MU,start=1)
summary(MU.mcmc)
(mean(y_male) + mean(y_female))/2 #compare to data
```



---
## Posterior summaries

```{r fig.height=4}
DELTA.mcmc <- mcmc(DELTA,start=1)
summary(DELTA.mcmc)
summary((2*DELTA)) #rescale as difference in group means
mean(y_male) - mean(y_female) #compare to data
```



---
## Posterior summaries

```{r fig.height=4}
SIGMA2.mcmc <- mcmc(SIGMA2,start=1)
summary(SIGMA2.mcmc)
```



---
## Diagnostics

```{r fig.height=4}
plot(MU.mcmc)
```



---
## Diagnostics

```{r fig.height=4}
autocorr.plot(MU.mcmc)
```



---
## Diagnostics

```{r fig.height=4}
plot(DELTA.mcmc)
```



---
## Diagnostics

```{r fig.height=4}
autocorr.plot(DELTA.mcmc)
```



---
## Diagnostics

```{r fig.height=4}
plot(SIGMA2.mcmc)
```



---
## Diagnostics

```{r fig.height=4}
autocorr.plot(SIGMA2.mcmc)
```



---
## Application to data

- Posterior probability that boys have larger average BMI than girls is 0.86!

- Posterior medians and 95% credible intervals for the group means are actually quite similar to the unpooled gender specific intervals from classical inference (do a t-test to confirm).
  ```{r}
#mean for boys
quantile((MU+DELTA),probs=c(0.025,0.5,0.975))

#mean for girls
quantile((MU-DELTA),probs=c(0.025,0.5,0.975))

#posterior probability boys have larger BMI than girls
mean(DELTA > 0)
```



---
## Application to data

- Let's look at a different sub-population. For older individuals $> 75$, we only have 8 male and 4 female.
  ```{r}
y_male <- bmi[Howell1$age > 75 & Howell1$male==1]
y_female <- bmi[Howell1$age > 75 & Howell1$male==0]
n_m <- length(y_male)
n_f <- length(y_female)
n_m
n_f
```



---
## Application to data

- A 95% confidence interval for the difference between genders in BMI (estimated as 0.24) is (-4.20,4.68).
  ```{r}
mean(y_male) - mean(y_female)
t.test(y_male,y_female)
```




---
## Application to data

- Let's apply the Bayesian model with these priors:
  + $\mu_0 = 18, \gamma_0 = 5$,
  + $\delta_0 = 0, \tau_0 = 3$,
  + $\nu_0 = 1, \sigma_0 = 5$.
  
```{r, cache=T, echo=F}
#priors
mu0 <- 18; gamma02 <- 5^2
delta0 <- 0; tau02 <- 3^2
nu0 <- 1; sigma02 <- 5^2

#starting values
mu <- (mean(y_male) + mean(y_female))/2
delta <- (mean(y_male) - mean(y_female))/2
#no need for starting values for sigma_squared, we can sample it first

MU <- DELTA <- SIGMA2 <- NULL
#set seed
set.seed(1234)

#set number of iterations and burn-in
n_iter <- 10000; burn_in <- 0.2*n_iter

##Gibbs sampler
for (s in 1:(n_iter+burn_in)) {
#update sigma2
sigma2 <- 1/rgamma(1,(nu0 + n_m + n_f)/2,
                   (nu0*sigma02 + sum((y_male-mu-delta)^2) + sum((y_female-mu+delta)^2))/2)

#update mu
gamma2n <- 1/(1/gamma02 + (n_m + n_f)/sigma2)
mun <- gamma2n*(mu0/gamma02 + sum(y_male-delta)/sigma2 + sum(y_female+delta)/sigma2)
mu <- rnorm(1,mun,sqrt(gamma2n))

#update delta
tau2n <- 1/(1/tau02 + (n_m+n_f)/sigma2)
deltan <- tau2n*(delta0/tau02 + sum(y_male-mu)/sigma2 - sum(y_female-mu)/sigma2)
delta <- rnorm(1,deltan,sqrt(tau2n))

#save parameter values
MU <- c(MU,mu); DELTA <- c(DELTA,delta); SIGMA2 <- c(SIGMA2,sigma2)
}
```

--

- The R code for running the sampler is suppressed here. Basically, just re-run the same Gibbs sampler from before on this new data.

--

- Using the results from the model, the posterior mean is 0.25 with 95% CI (-3.45, 3.88).
  ```{r}
mean((DELTA*2))
quantile((DELTA*2),probs=c(0.025,0.5,0.975))
```




---
## Application to data

- The width of this interval is smaller than that of the 95% confidence interval from before. 

--

- In a way, precision has been improved by borrowing of information across the groups. Of course the prior is important here given the sample sizes.




---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




