---
title: "Hierarchical models III"
author: "Dr. Olanrewaju Michael Akande"
date: "Oct 22, 2019"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
#library(lattice)
#library(dplyr)
#library(ggplot2)
library(lme4)
library(arm)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```

## Announcements
  
- Reminder: submissions for final project proposals will not be graded.

- Full instructions for the proposals will be posted by the end of the week.

- Updated dates for final project presentations

## Outline

- Questions from last lecture

- Bernoulli versus binomial outcomes

- Multilevel logistic models

- Bayesian approach to multilevel logistic models using the `brms` package


---
## Bernoulli versus binomial outcomes

- Before we get into multilevel logistic models, let's review logistic regression models for Bernoulli versus binomial outcomes.

--

- Recall from the summer online review materials that if $Y \sim \textrm{Bin}(n,p)$ (that is, $Y$ is a random variable that follows a binomial distribution with parameters $n$ and $p$), then $Y$ follows a $\textrm{Bernoulli}(p)$ distribution  when $n = 1$.

--

- This also implies that if $Z_1, \ldots, Z_n \sim \textrm{Bernoulli}(p)$, then $Y = \sum_i^n Z_i \sim \textrm{Bin}(n.p)$.

--

- That is, the sum of $n$ iid $\textrm{Bernoulli}(p)$ random variables gives a random variable with the $\textrm{Bin}(n,p)$ distribution.

--

- The logistic regression model can be used either for Bernoulli data (as we have done so far) or for data summarized as binomial counts (that is, aggregated counts).

--

- This is what you have for one part of team project 2.


---
## Bernoulli versus binomial outcomes

Normally, for individual-level data, we would have

.small[
```{r echo=F}
set.seed(10); n <- 100; y <- sample(c(0,1),n,replace=T); x <- factor(sample(c(1:3),n,replace=T))
Data <- data.frame(y=y,x=x); head(Data)
```
]

.small[
```{r}
M1 <- glm(y~x,data=Data,family=binomial); summary(M1)
```
]


---
## Bernoulli versus binomial outcomes

But we could also do the following with the aggregate level data instead.
.small[
```{r echo=F}
Data_agg <- aggregate(Data$y,by=list(x),FUN=sum); colnames(Data_agg) <- c("x","success")
Data_agg$n <- aggregate(Data$y,by=list(x),FUN=length)$x; Data_agg
```
]

.small[
```{r}
M2 <- glm(cbind(success,n-success)~x,data=Data_agg,family=binomial); summary(M2)
```
]

--

Same results overall! Deviance and AIC are different because of the different likelihood functions. Note that some glm functions use `n` instead of `n-success`.


---
class: center, middle

# Multilevel logistic models



---
## Multilevel linear models

- The same idea and approach used to build multilevel models for normal data can be used to build multilevel logistic (and probit) models for binary outcomes.

--

- Recall that for a .hlight[varying-intercepts linear model] with one individual-level predictor, we have
.block[
.small[
$$
\begin{split}
y_{ij} & = \beta_{0} + \gamma_{0j} + \beta_1 x_{1ij} + \epsilon_{ij}; \ \ \ i = 1, \ldots, n_j; \ \ \ j = 1, \ldots, J; \\
\epsilon_{ij} & \sim N(0, \sigma^2); \\
\gamma_{0j} & \sim N(0, \tau_0^2)
\end{split}
$$
]
]

  where $x_{1ij}$ can be replaced with $x_{1j}$ for a group-level predictor.

--

- This model is a compromise between complete pooling across groups of a grouping variable, such as counties in the radon example for last class (that is, same intercept for each county), and no pooling (estimating a separate intercept for each county without borrowing information).

--

- The degree of pooling is determined by the amount of information within and between groups.


---
## Multilevel logistic models

- We can use the same idea to build a .hlight[varying-intercepts logistic model].

- That is,
.block[
.small[
$$
\begin{split}
y_{ij} | x_{ij} & \sim \textrm{Bernoulli}(\pi_{ij}); \ \ \ i = 1, \ldots, n_j; \ \ \ j = 1, \ldots, J; \\
\textrm{log}\left(\dfrac{\pi_{ij}}{1-\pi_{ij}}\right) & = \beta_{0} + \gamma_{0j} + \beta_1 x_{1ij}; \\
\gamma_{0j} & \sim N(0, \sigma_0^2)
\end{split}
$$
]
]

  where once again, $x_{1ij}$ is an individual-level predictor which can be replaced with $x_{1j}$ for a group-level predictor.


--

- The Gelman and Hill book uses the following notation instead
.block[
.small[
$$
\begin{split}
y_i | x_i & \sim \textrm{Bernoulli}(\pi_i); \ \ \ i = 1, \ldots, n; \ \ \ j = 1, \ldots, J; \\
\textrm{log}\left(\dfrac{\pi_i}{1-\pi_i}\right) & = \beta_{0} + \gamma_{0j[i]} + \beta_1 x_{i1}; \\
\gamma_{0j} & \sim N(0, \sigma_0^2).
\end{split}
$$
]
]


--

- I will use this notation for the rest of today's class.


---
## Varying-intercepts logistic model

Inverse logit functions for varying-intercepts logistic models.

```{r fig.height=4,echo=F,warning=F}
curve(invlogit((0.3*x)),xlim=c(0,5),ylim=c(0,1),col="blue4",ylab="probabilities")
curve(invlogit((1+0.3*x)),xlim=c(0,5),ylim=c(0,1),col="red4",add=T)
curve(invlogit((1.5+0.3*x)),xlim=c(0,5),ylim=c(0,1),col="green4",add=T)
curve(invlogit((3+0.3*x)),xlim=c(0,5),ylim=c(0,1),col="orange4",add=T)
```


---
## Multilevel logistic models

- It is easy to extend this model to allow for .hlight[varying-slopes] instead or both .hlight[varying-intercepts and varying-slopes] just like we had for multilevel linear models.

--

- The interpretations of the fixed effect(s) in multilevel logistic models follow directly from what we had for the standard logistic models, that is, log-odds, odds and odds ratios.

--

- The only difference now is the hierarchy in our data which allows us to borrow information across groups.

--

- One way to think about this is that we expect odds and odd-ratios to be more similar for observations within the same group, but we allow for some similarity across groups via partial pooling.


---
## Varying-slopes logistic model

Inverse logit functions for varying-slopes logistic models.

```{r fig.height=4,echo=F,warning=F}
curve(invlogit((0.3*x)),xlim=c(0,5),ylim=c(0,1),col="blue4",ylab="probabilities")
curve(invlogit((0.5*x)),xlim=c(0,5),ylim=c(0,1),col="red4",add=T)
curve(invlogit((0.7*x)),xlim=c(0,5),ylim=c(0,1),col="green4",add=T)
curve(invlogit((x)),xlim=c(0,5),ylim=c(0,1),col="orange4",add=T)
```


---
## Varying-intercepts, varying-slopes logistic model

Inverse logit functions for varying-intercepts, varying-slopes logistic models.

```{r fig.height=4,echo=F,warning=F}
curve(invlogit((0.3*x)),xlim=c(0,5),ylim=c(0,1),col="blue4",ylab="probabilities")
curve(invlogit((1+0.5*x)),xlim=c(0,5),ylim=c(0,1),col="red4",add=T)
curve(invlogit((1.5-0.7*x)),xlim=c(0,5),ylim=c(0,1),col="green4",add=T)
curve(invlogit((3-x)),xlim=c(0,5),ylim=c(0,1),col="orange4",add=T)
```


---
## 1988 elections analysis

- To illustrate how to fit and interpret the results of multilevel logistic models, we will use the election polls example from the Gelman and Hill textbook.

--

- National opinion polls are conducted by a variety of organizations (e.g., media, polling organizations, campaigns) leading up to elections.

--

- While many of the best opinion polls are conducted at a national level, it can also be often interesting to estimate voting opinions and preferences at the state or even local level.

--

- Well-designed polls are generally based on national random samples with corrections for nonresponse based on a variety of demographic factors (e.g., sex, ethnicity, race, age, education).



---
## 1988 elections analysis

- The data is from CBS News surveys conducted during the week before the 1988 election.

--

- Respondents were asked about their preferences for either the Republican candidate (Bush Sr.) or the Democratic candidate (Dukakis). 

--

- Both voting turnout and preferences often depend on a complex combination of demographic factors.

--

- In our example dataset, we have demographic factors such as biological sex, race, age, education, which we may all want to look at by state, resulting in $2 \times 2 \times 4 \times 4 \times 51 = 3264$ potential categories of respondents.

--

- Clearly, without a very large survey (most political survey poll around 1000 people), we will need to make assumptions in order to even obtain estimates in each category.

--

- We usually cannot include all interactions; we should therefore select those to explore (through EDA and background knowledge).



---
## 1988 elections analysis

The dataset includes 2193 observations from one of eight surveys (the most recent CBS News survey right before the election) in the original full data.

.small[
Variable    | Description
:------------- | :------------
org | cbsnyt = CBS/NYT
bush | 1 = preference for Bush Sr., 0 = otherwise
state | 1-51: 50 states including DC (number 9)
edu | education: 1=No HS, 2=HS, 3=Some College, 4=College Grad
age | 1=18-29, 2=30-44, 3=45-64, 4=65+
female | 1=female, 0=male
black | 1=black, 0=otherwise
region | 1=NE, 2=S, 3=N, 4=W, 5=DC
v_prev | average Republican vote share in the three previous elections (adjusted for home-state and home-region effects in the previous elections)
]

The data is in the file `polls_subset.txt` on Sakai.


---
## 1988 elections analysis

```{r}
###### Load the data
polls_subset <- read.table("data/polls_subset.txt",header=TRUE)

polls_subset$v_prev <- polls_subset$v_prev*100 #rescale 
polls_subset$region_label <- factor(polls_subset$region,levels=1:5,
                                    labels=c("NE","S","N","W","DC"))
#we consider DC as a separate region due to its distinctive voting patterns
polls_subset$edu_label <- factor(polls_subset$edu,levels=1:4,
                                 labels=c("No HS","HS","Some College","College Grad"))
polls_subset$age_label <- factor(polls_subset$age,levels=1:4,
                                 labels=c("18-29","30-44","45-64","65+"))
#the data includes states but without the names, which we will need, 
#so let's grab that from R datasets
data(state) 
#"state" is an R data file (type ?state from the R command window for info)
state.abb #does not include DC, so we will create ours
#In the polls data, DC is the 9th "state" in alphabetical order
state_abbr <- c (state.abb[1:8], "DC", state.abb[9:50])
polls_subset$state_label <- factor(polls_subset$state,levels=1:51,labels=state_abbr)
rm(list = ls(pattern = "state")) #remove unnecessary values in the environment
```



---
## 1988 elections analysis

```{r}
###### View properties of the data  
head(polls_subset)
dim(polls_subset)
```



---
## 1988 elections analysis

```{r}
###### View properties of the data  
str(polls_subset)
```


---
## 1988 elections analysis

- I will not do any substantial EDA here.

--

- I expect you to be able to do this yourself!

--

- Let's just take a look at the amount of data we have for "bush" and the age:edu interaction

  ```{r}
###### Exploratory data analysis
table(polls_subset$bush) #well split by the two values
table(polls_subset$edu,polls_subset$age)
```


---
## 1988 elections analysis

- As a start, we will consider a simple model with fixed effects of race and sex and a random effect for state (50 states + the District of Columbia).
.block[
.small[
$$
\begin{split}
\textrm{bush}_i | \boldsymbol{x}_i & \sim \textrm{Bernoulli}(\pi_i); \ \ \ i = 1, \ldots, n; \ \ \ j = 1, \ldots, J=51; \\
\textrm{log}\left(\dfrac{\pi_i}{1-\pi_i}\right) & = \beta_{0} + \gamma_{0j[i]} + \beta_1 \textrm{female}_{i} + \beta_2 \textrm{black}_{i}; \\
\gamma_{0j} & \sim N(0, \sigma_{\textrm{state}}^2).
\end{split}
$$
]
]

--

- We can also write
.block[
.small[
$$
\begin{split}
\textrm{bush}_i | \boldsymbol{x}_i & \sim \textrm{Bernoulli}(\pi_i); \ \ \ i = 1, \ldots, n; \ \ \ j = 1, \ldots, J=51; \\
\textrm{log}\left(\dfrac{\pi_i}{1-\pi_i}\right) & = \beta_{0} + \gamma_{0j[i]}^{\textrm{state}} + \beta_{\textrm{female}} \textrm{female}_{i} + \beta_{\textrm{black}} \textrm{black}_{i}; \\
\gamma_{0j} & \sim N(0, \sigma_{\textrm{state}}^2).
\end{split}
$$
]
]

--

- In `R`, we have
  ```{r results = 'hide'}
library(lme4)
model1 <- glmer(bush ~ black+female+(1|state_label),family=binomial(link="logit"),
data=polls_subset)
summary(model1)
```


---
## 1988 elections analysis
```{r echo=F}
summary(model1)
```


---
## 1988 elections analysis

- Looks like we dropped some NAs.
  ```{r}
c(sum(complete.cases(polls_subset)),sum(!complete.cases(polls_subset)))
```

--

- Not ideal; we'll learn about methods for dealing with missing data soon.

--

- Interpretation of results:
  + For a fixed state, a non-black male respondent has odds of $e^{0.45}=1.57$ of supporting Bush.
  + For a fixed state and sex, a black respondent as $e^{-1.74}=0.18$ times (an 82% decrease)  the odds of supporting Bush as a non-black respondent; you are much less likely to support Bush if your race is black compared to being non-black.
  + For a given state and race, a female respondent has $e^{-0.10}=0.91$ (a 9% decrease)  times the odds of supporting Bush as a male respondent. However, this effect is not actually statistically significant!



---
## 1988 elections analysis

- The state-level standard deviation is estimated at 0.41, so that the states do vary some, but not so much.

--

-   <div class="question">
We no longer have a term for residual standard deviation (residual standard error). Why is that?
</div>

--


- I expect that you will be able to interpret the corresponding confidence intervals.
  ```{r echo=F}
confint(model1)
```


---
## 1988 elections analysis

- Let’s fit a more sophisticated model that includes other relevant survey factors, such as 
  + region
  + prior vote history (note that this is a state-level predictor),
  + age, education, and the interaction between them.

--

- In `R`, we have
  ```{r results = 'hide'}
model2 <- glmer(bush ~ black + female + v_prev + edu_label:age_label +
                (1|state_label) + (1|region_label),
                family=binomial(link="logit"),data=polls_subset)
```

--

-   <div class="question">
Why do we have a rank deficient model?
</div>



---
## 1988 elections analysis

- Also, it looks like we have a convergence issue. This can happen when dealing with multilevel models. We have so many parameters to estimate from the interaction terms `edu_label:age_label` (16 actually), and it looks like that's causing a problem. 

--

- Could be that we have too many $\textrm{bush}_i = 1$ or $0$ values for certain combinations. You should check!

--

- Let's treat those as random effects instead. That is,
.block[
.small[
$$
\begin{split}
\textrm{logit}\left(\Pr[\textrm{bush}_i = 1] \right) & = \beta_{0} + \gamma_{0m[i]}^{\textrm{region}} + \gamma_{0j[i]}^{\textrm{state}} + \gamma_{0k[i],l[i]}^{\textrm{age.edu}} \\
& + \beta_{\textrm{f}} \textrm{female}_{i} + \beta_{\textrm{b}} \textrm{black}_{i} + \beta_{\textrm{v_prev}} \textrm{v_prev}_{j[i]}; \\
\gamma_{0m} & \sim N(0, \sigma_{\textrm{region}}^2), \ \ \ \gamma_{0j} \sim N(0, \sigma_{\textrm{state}}^2), \ \ \ \gamma_{0k,l} \sim N(0, \sigma_{\textrm{age.edu}}^2).
\end{split}
$$
]
]

- In `R`, we have
  ```{r results = 'hide'}
model3 <- glmer(bush ~ black + female + v_prev + (1|state_label) + (1|region_label) + 
               (1|edu_label:age_label), family=binomial(link="logit"),data=polls_subset)
```

--

- This seems to run fine; we are able to borrow information which will help.


---
## 1988 elections analysis

.small[
```{r echo=F}
summary(model3)
```
]



---
## 1988 elections analysis

- Remember that in the first model, the state-level standard deviation was estimated as 0.41. Looks like we are now able to separate that (for the most part) into state and region effects.

--

- Interpretation of results:
  + For a fixed state, education and age bracket, a non-black male respondent with zero prior average Republican vote share, has odds of $e^{-3.51}=0.03$ of supporting Bush (no one really has 0 value for `v_prev`).
  + For a fixed state, sex, education level, age bracket and zero prior average Republican vote share, a black respondent has $e^{-1.75}=0.17$ times (an 83% decrease) the odds of supporting Bush as a non-black respondent, which is about the same as before.
  + For each percentage point increase in prior average Republican vote share, residents of a given state, race, sex, education level  age bracket have $e^{0.07}=1.07$ times the odds of supporting Bush.


---
## 1988 elections analysis: Bayesian approach

- Due to the number of categories, the inference in the frequentist model is not entirely reliable as
  + it does not fully account for uncertainty in the estimated variance parameters, and 
  + it uses an approximation for inference. 
  
--

- We can fit the model under the Bayesian paradigm in the `brms` package, using mildly informative priors and quantify uncertainty based on the posterior samples.

--

- In-class analysis: move to the R script [here](https://ids-702-f19.github.io/Course-Website/slides/lec-slides/Elections88.R)

--

- Windows users: install Rtools for windows, then the `rstan` package in R. Mac users: install Xcode, Try `xcode-select --install` from the Terminal app or the Terminal tab of the RStudio console, then install the `rstan` package in R.




















