---
title: "Time series models"
author: "Dr. Olanrewaju Michael Akande"
date: "Nov 14, 2019"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
library(lattice)
#library(dplyr)
library(ggplot2)
#library(arm)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```


## Announcements

- Final lab tomorrow.

- Survey on slots for final project presentation over the next few days.

- Guidelines for final project presentations next week.

## Outline

- Introduction

- Stationarity

- AR models

- MA models

- ARMA, ARIMA??



---
## Introduction

- When data are ordered in time, responses and errors from one period may influence responses and errors from another period.

--

- For example, it is reasonable to expect unemployment rate in a month to be correlated with unemployment rate in previous month(s).

--

- Another example: weather events in current time period may depend on weather events in previous time period.

--

- These are called .hlight[time series] data.

--

- Correlation due to time is called serial correlation or autocorrelation.

--

- We will only scratch the surface in this course.


---
## Goals of time series analysis

- .hlight[Forecasting outcomes]

--

  + Given a series of outcomes ordered in time, predict the values of the outcomes in the future.

--
  
  + Examples:
      - forecasting future price of oil given historical oil prices.
      - predicting future price of a particular stock price given past prices of the same stock.

--

  + When forecasting, it is important to also report an interval estimate to incorporate uncertainty about future values.
  

---
## Goals of time series analysis

- .hlight[Forecasting outcomes]
  
--

  + Forecasting outcomes using predictors may involve building a model for the predictors as well, since we can't observe them in the future.
  
--

  + For example, predicting inflation rate given employment rate requires estimating future values for the employment rate as well.
  
--

- .hlight[Learning relationships with data ordered in time].
  
--

  + How are outcomes correlated over time? Are there periodic relationships in outcomes?
  
--

  + Regressions of outcomes on predictors, accounting for correlated errors due to time series.
  

---
## Motivating example: FTSE 100

- The FTSE (Financial Times Stock Exchange) 100 Index is a share index of the 100 companies listed on the London Stock Exchange with the highest market capitalization.
  
--

- A share index is essentially a form of weighted average of prices of selected stocks.
  
--

- To motivate our discussions on time series, let's look at data for FTSE 100 returns in 2018.
  ```{r}
ftse100 <- read.csv("data/ftse2018.csv", header = T)
head(ftse100)
```

- Can we forecast closing prices for the next five days from 11/7/2018?

---
## Motivating example: FTSE 100

Notice that the data go from latest to earliest date, so let's invert the order of the rows to make the time series increasing in date.
  
--

```{r}
ftse100 <- ftse100[nrow(ftse100):1,]
dim(ftse100)
head(ftse100)
```

---
## Motivating example: FTSE 100

Plot the closing prices to see what a simple time series data looks like.
```{r fig.height=2.8}
tsClose <- ts(ftse100$Close); ts.plot(tsClose,col="red3")
```
  
--

- It is reasonable to expect closing prices for a particular day to be correlated with closing prices for previous days. 
  
--

- How many of the previous days? We will have to investigate!


---
## Motivating example: Sunspots and melanoma

- We will revisit that data but let's look at different example, where we also have a predictor.
  
--

- Incidence of melanoma (skin cancer) may be related to solar radiation.
  
--

- Annual data from Connecticut tumor registry on age adjusted melanoma incidence rates (per 100000 people). Treat these rates as without error.
  
--

- We also have annual data on relative sunspot (dark spots on the sun caused by intense magnetic activity) activity.
  
--

- Data go from 1936 to 1972.


---
## Motivating example: Sunspots and melanoma

```{r fig.height=3}
cancersun <- read.csv("data/melanoma.csv", header = T)
names(cancersun) = c("year", "melanoma", "sunspot")
str(cancersun)
head(cancersun)
```


---
## Motivating example: Sunspots and melanoma

```{r fig.height=3}
ggplot(cancersun, aes(x=sunspot, y=melanoma)) +
  geom_point(alpha = .5,colour="blue4") +
  geom_smooth(method="lm",col="red3") +
  labs(title="Melanoma Incidence Rate vs Sunspots")
```

Weak positive (maybe!) relationship between them.


---
## Motivating example: Sunspots and melanoma

Let's look at melanoma incidence rate in time
```{r fig.height=3}
tsMelanoma <- ts(cancersun$melanoma); ts.plot(tsMelanoma,col="blue4")
```

Trend in time, some of which we might be able to explain using `sunspots`.


---
## Motivating example: Sunspots and melanoma

Let's fit a linear model to the relationship between the two variables.
```{r fig.height=2.8}
regmelanoma = lm(melanoma ~ sunspot, data = cancersun)
ggplot(cancersun, aes(x=sunspot, y=regmelanoma$residual)) +
  geom_point(alpha = .5,colour="blue4") +
  geom_smooth(method="lm",col="red3") + labs(title="Residuals vs Sunspots")
```

Residuals look fine here.

---
## Motivating example: Sunspots and melanoma

Let's plot the residuals versus year.
```{r fig.height=2.8}
ggplot(cancersun, aes(x=year, y=regmelanoma$residual)) +
  geom_point(alpha = .5,colour="blue4") +
  geom_smooth(method="lm",col="red3") + labs(title="Residuals vs Year")
```

Huge trend! What to do???


---
## Stationarity

- The most common time series models usually assume what is known as .hlight[stationarity].
  
--

- .hlight[Stationarity] of a time series process means that the marginal distribution of any part of the series does not depend on time.
  
--

- Basically, the locations in time themselves does not matter in the marginal and joint distributions; however, the differences in locations, that is, the lags, do matter!
  
--

- Put a different way, a stationary time series is one whose properties do not depend on the particular time at which the series is observed. 
  
--

- Examples of non stationary series:
  + Steadily increasing trend (like the melanoma example).
  + Known seasonal trends, like increase in sales before Christmas.
  + Break in trend due to some external event.
  

---
## Stationarity

- Denote the times series for the outcome by $y_t$.
  
--

- Stationarity $\Rightarrow$
  + $\mathbb{Pr}(y_t)$ is the same for all $t$, 
--

  + $\mathbb{Pr}(y_t,y_{t+1})$ is the same for all $t$,  
  and so on...
  
--

- .hlight[Weak stationarity] requires that only marginal moments, that is, means, variances and covariances are the same.
  
--

- Stationarity $\Rightarrow$ weak stationarity, but the converse need not hold.
  
--

- For a normal distribution, the mean and variance completely characterizes the distribution, so that stationarity and weak stationarity will be equivalent. 
  
--

- Why does that matter? When dealing with linear models, what distribution do we assume?...
  
---
## Popular stationary models

- We will mainly focus on two types of stationary time series models:
  
--

  + Autoregressive models (AR models)
      - Value of outcome at time $t$ is correlated with value at previous times.
        
--

  + Moving average models (MA models)
      - Value of outcome at time $t$ is correlated with value of prediction errors at previous times.
        
--

  
- Side note: autoregressive moving average models (ARMA models)
    - Combination of AR and MA.
      
--

- There are many more types of time series models (see STA 642/942).


---
## Autocorrelation

- .hlight[Autocorrelation] (serial correlation) measures the strength of the linear relationship between $y_t$ and its lagged values.
  
--

- The lag $k$ autocorrelation $\rho_k$ measures the correlation in outcomes at time $t$ and at time $t-k$, where $k$ indicates how far back to go; $k$ is called a lag.
  
--

- Recall that for two random variables $X$ and $Y$, we have
.block[
.small[
$$\mathbb{Corr}[X,Y] = \dfrac{\mathbb{E}[(X-\mu_X)(Y-\mu_Y)]}{\sqrt{\mathbb{V}[X]\mathbb{V}[Y]}}.$$
]
]
  
--

- The sample autocorrelation $r_k$ can be calculated using
.block[
.small[
$$r_k = \dfrac{\sum_{t=k+1}^T(y_t - \bar{y})(y_{t-k} - \bar{y})}{\sum_{t=1}^T(y_t - \bar{y})^2}.$$
]
]
  
--

-  .question[Why do we only have an estimate of the variance of y at time t in the denominator?] Think stationarity...


---
## Partial autocorrelation

- Autocorrelation $\rho_k$ (and the sample version $r_k$) between $y_t$ and $y_{t-k}$ will also include the linear relationships between $y_t$  and each of $y_{t-1}, y_{t-2}, \ldots, y_{t-k+1}$.
  
--

- As you will see, we will need to be able to assess the correlation between $y_t$ and $y_{t-k}$ without interference from the other lags.
  
--

- .hlight[Partial autocorrelation] lets us do just that. It is the autocorrelation between $y_t$ and $y_{t-k}$, with all the linear relationships between $y_t$  and each of $y_{t-1}, y_{t-2}, \ldots, y_{t-k+1}$ removed.
  
--

- In `R`, use `acf` to compute and plot autocorrelations and `pacf` to compute and plot partial autocorrelations.


---
## AR models

- The most common time series model is called the .hlight[autoregressive (AR)] model.
  
--

- When only one lag matters, the zero-mean AR(1) model is
.block[
.small[
$$y_t = \phi y_{t-1} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- With a non-zero mean, we have
.block[
.small[
$$y_t = \mu + \phi y_{t-1} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- When the mean is non-zero, we can choose to de-mean (mean-center) the series and model that instead.
  
--

- In both cases, for the AR(1) we basically have a linear regression where the value of the outcome at time $t$ depends on value of outcome at time $t-1$.
  
--

- $\phi$ is the autocorrelation.


---
## AR models

- For the zero-mean AR(1) model,  

--

  + $|\phi|<1$ represents stationary time series.
    
--

  + $\phi=1$ is a random walk.
    
--

  + $|\phi|>1$ implies non-stationary, "explosive" models.
    
--

  
- A stationary AR(1) series varies around its mean, randomly wandering off away from the mean in response to the "input" values of the random $\epsilon_t$ series, but always returning to near the mean, and never "exploding" away for more than a short time.
  
--

- AR(1) series with $0<\phi<1$ represent short-term, positive correlations that would damp out exponentially if $\epsilon_t$ were zero.
  
--

- Negative values of $\phi$ represent short-term, negative correlations.


---
## AR models

- Let's explore what AR(1) models look like via simulations.
  
--

- Move to the R script [here](https://ids-702-f19.github.io/Course-Website/slides/lec-slides/TS_simulations.R).

--

- Note that
  
--

  + autocorrelations decay steadily with lags.
    
--

  + partial autocorrelations go to zero after lag p.


---
## AR models

- For a zero mean AR(p) model, we have
.block[
.small[
$$y_t = \sum_{k=1}^{p} \phi_k y_{t-k} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- So that for a non-zero mean AR(p) model, we have
.block[
.small[
$$y_t = \mu + \sum_{k=1}^{p} \phi_k y_{t-k} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- AR(p) models are capable of adequately representing a wide range of observed behaviors in time series for large enough $p$.


---
## AR models: how many lags?

- Several ways to decide how many lags to include.
  
--

- Use graphical techniques
  + Look at partial autocorrelation plots.
    
--

  + Set $p$ at lag where correlations become small enough not to be important.
  
--

- Use a model selection criterion like BIC.
  
--

- See section 8.6 of the assigned readings.
  
--

- Sometimes in time series data, the partial autocorrelations are small even at lag 1.
  
--

- In this case, it can be reasonable to skip autoregressive models and just use usual linear regression modeling approaches.


---
## What if the series is not stationary?

- Sometimes transformations can make stationarity a reasonable assumption.
  
--

- Differencing (subtract lagged values from outcome at time $t$) also often help; changes over time are more likely to be stationary than the raw values.
  
--

- Including predictors can also help as we will see later with the melanoma example.
  
--

- There are other models for non stationary time series.


---
## AR(p): including predictors

- We also might want to account for serial correlation in regression modeling.
  
--

- Linear regression assumes independent errors across individuals.
  
--

- As we have already seen with the melanoma example, this may not be reasonable with time series data.
  
--

- With a single predictor $x_t$, we have
.block[
.small[
$$y_t = \mu + \sum_{k=1}^{p} \phi_k y_{t-k} + x_t + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- That is, the value of outcome at time $t$ depends on value of outcome at time $t-1, t-2, \ldots, t-k$, but also on the predictor $x$ at time $t$.
  
--

- Easy to extend the model to multiple predictors.


---
## Model assumptions: stationarity

- Coefficients and regression variance do not change with time.
  
--

  + Apart from changes in explanatory variables, the behavior of the time series is the same at different segments of time.
    
--

  + Generally, no predictable patterns in the long term
    
--

- Diagnostics: check if patterns in residuals are similar across time.
  
--

- Tests:
  + Ljung-Box
  + Augmented Dickey–Fuller (ADF)
  + Kwiatkowski-Phillips-Schmidt-Shin (KPSS)
  
--

- Remedies:
  + Sometimes transformations (e.g., using logs) can make stationarity more reasonable.
  + Use time series models that allows drift.
  
  
---
## Model assumptions: others

- Other assumptions
  1. Linearity
  
  2. Independence of errors
  
  3. Equal variance

  4. Normality
  
--

- Diagnose using the same methods we used for linear regression.
  
--

- Remedies include transformations and model changes as we had before.


---
## MA models

- The zero-mean MA(1) model is
.block[
.small[
$$y_t = \phi \epsilon_{t-1} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- With a non-zero mean, we have
.block[
.small[
$$y_t = \mu + \phi \epsilon_{t-1} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- The value of the outcome at time $t$ depends on the value of the deviation from the mean (the error term) at time $t-1$.
  
--

- For a zero mean MA(p) model, we have
.block[
.small[
$$y_t = \sum_{k=1}^{p} \phi_k \epsilon_{t-k} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]
  
--

- So that for a non-zero mean MA(p) model, we have
.block[
.small[
$$y_t = \mu + \sum_{k=1}^{p} \phi_k \epsilon_{t-k} + \epsilon_t; \ \ \epsilon_t \sim N(0, \sigma^2).$$
]
]


---
## MA models

- Let's explore what MA(1) models looks like via simulations. Move back to the same R script.

--

- Note that
  
--

  + Autocorrelations die off almost immediately after lag 1.
    
--

  + In MA(p) model, autocorrelations (mostly!) die off after lag $p$. May not be exact since autocorrelation measures correlation between the actual outcome at different time points.
    
--

  + Partial autocorrelations are not particularly useful.
  
<!-- - Contrast to AR(p) models -->
<!--   + Autocorrelations tend to decrease over time smoothly. -->
<!--   + Partial autocorrelations die off after lag $p$. -->
    
--

- It is possible to write any stationary AR(p) model as an $\textrm{MA}(\infty)$ model. The reverse result holds for some constraints on the MA parameters. See the reading material.
  
  
---
## Deciding models?

- Use autocorrelations and partial autocorrelations to help decide model.
  
--

- Steady decay on autocorrelations often implies AR.
  
--

- Non zero autocorrelations before lag $p$ and zero after lag $p$ often implies MA.
  
--

- Sometimes use both AR and MA error structure, called an .hlight[ARMA] model.
  
--

- Whenever we take differences in $y$ values to ensure stationarity before fitting ARMA models, we have .hlight[ARIMA] models.





