---
title: "Time series models (cont'd)"
author: "Dr. Olanrewaju Michael Akande"
date: "Nov 19, 2019"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
#library(tidyverse)
#library(magick)
library(knitr)
library(kableExtra)
library(lattice)
#library(dplyr)
library(ggplot2)
#library(arm)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  #show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
knitr::opts_chunk$set(fig.height = 2.65,dpi =300,fig.align='center',fig.show='hold',size='footnotesize',small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}

knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar =  c(4, 4, 1.5, 1.5)) 
})
```


## Announcements

- Survey on slots for final project presentation out after class.

- I will send out presentation slots on Thursday.

- Final project presentations:
  + Each student will give a brief (5 minute) presentation of their findings in class.
  + Presentations will be graded for:
      - Preparedness
      - Clarity
      - Time management
      - Understanding of the data/project
  + Each student's presentation must be submitted on gradescope by 8:00am on the day they present.



---
class: center, middle

# In-class analysis: move to the R script [here](https://ids-702-f19.github.io/Course-Website/slides/lec-slides/TS_examples.R)




