---
title: "STA 360/602L: Module 1.1"
subtitle: "Building blocks of Bayesian inference"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```



## Building blocks of Bayesian inference

- Generally (unless otherwise stated), in this course, we will use the following notation. Let

--

  + $\mathcal{Y}$ be the .hlight[sample space];

--

  + $y$ be the .hlight[observed data];

--

  + $\Theta$ be the .hlight[parameter space]; and

--

  + $\theta$ be the .hlight[parameter of interest].
  
--

- More to come later.


---
## Frequentist inference

- Given data $y$, estimate the population parameter $\theta$.

--

- How to estimate $\theta$ under the frequentist paradigm?
  + Maximum likelihood estimate (MLE)
  
  + Method of moments
  
  + and so on...
  
--

- Frequentist ML estimation finds the one value of $\theta$ that maximizes the likelihood.

--

- Typically uses large sample (asymptotic) theory to obtain confidence intervals and do hypothesis testing.


---
## What are Bayesian methods?

- .hlight[Bayesian methods] are data analysis tools derived from the principles of Bayesian inference and provide

--

  + parameter estimates with good statistical properties;
  
--

  + parsimonious descriptions of observed data;
  
--

  + predictions for missing data and forecasts of future data; and
  
--

  + a computational framework for model estimation, selection,
and validation.


---
## Bayes' theorem - basic conditional probability

- Let's take a step back and quickly review the basic form of Bayes' theorem.

--

- Suppose there are some events $A$ and B having probabilities $\Pr(A)$ and $\Pr(B)$.

--

- Bayes' rule gives the relationship between the marginal probabilities of A and B and the conditional probabilities.

--

- In particular, the basic form of .hlight[Bayes' rule] or .hlight[Bayes' theorem] is
.block[
.small[
$$\Pr(A | B) = \frac{\Pr(A \ \textrm{and} \ B)}{\Pr(B)} = \frac{\Pr(B|A)\Pr(A)}{\Pr(B)}$$
]
]

--
$\Pr(A)$ = marginal probability of event $A$, $\Pr(B | A)$ = conditional probability of event $B$ given event $A$, and so on.


---
## Building blocks of Bayesian inference

- Now, to a slightly more complicated version of Bayes' rule. First,

  1. For each $\theta \in \Theta$, specify a .hlight[prior distribution] $p(\theta)$ or $\pi(\theta)$, describing our beliefs about $\theta$ being the true population parameter.

--

  2. For each $\theta \in \Theta$ and $y \in \mathcal{Y}$, specify a .hlight[sampling distribution] $p(y|\theta)$, describing our belief that the data we see $y$ is the outcome of a study with true parameter $\theta$. $p(y|\theta)$ gets us the .hlight[likelihood] $L(\theta|y)$.

--

  3. After observing the data $y$, for each $\theta \in \Theta$, update the prior distribution to a .hlight[posterior distribution] $p(\theta | y)$ or $\pi(\theta | y)$, describing our "updated" belief about $\theta$ being the true population parameter.

--

- Now, how do we get from Step 1 to 3? .hlight[Bayes' rule]!
.block[
.small[
$$p(\theta | y) = \frac{p(\theta)p(y|\theta)}{\int_{\Theta}p(\tilde{\theta})p(y| \tilde{\theta}) \textrm{d}\tilde{\theta}} = \frac{p(\theta)p(y|\theta)}{p(y)}$$
]
]

  We will use this over and over throughout the course!

  
  
---
## Notes on prior distributions

Many types of priors may be of interest. These may

  + represent our own beliefs; 

--

  + represent beliefs of a variety of people with differing prior opinions; or

--

  + assign probability more or less evenly over a large region of the parameter space.
  
--

  + and so on...
  

---
## Notes on prior distributions

- .hlight[Subjective Bayes]: a prior should accurately quantify some individual's beliefs about $\theta$.

--

- .hlight[Objective Bayes]: the prior should be chosen to produce a procedure with "good" operating characteristics without including subjective prior knowledge.

--

- .hlight[Weakly informative]: prior centered in a plausible region but not overly-informative, as there is a tendency to be over
confident about one's beliefs.


---
## Notes on prior distributions

- The prior quantifies your initial uncertainty in $\theta$ before you observe new data (new information) - this may be necessarily subjective & summarize experience in a field or
prior research.

--

- Even if the prior is not "perfect", placing higher probability in a ballpark of the truth leads to better performance.

--

- Hence, it is very seldom the case that a weakly informative prior is not preferred over no prior.

--

- One (very important) role of the prior is to stabilize estimates in the presence of limited data.

  
  
---
## Simple example - estimating a population proportion

- Suppose $\theta \in (0,1)$ is the population proportion of individuals with diabetes in the US.

--

- A prior distribution for $\theta$ would correspond to some distribution that distributes probability across $(0, 1)$.

--

- A very precise prior corresponding to abundant prior knowledge would be concentrated tightly in a small sub-interval of $(0, 1)$.

--

- A vague prior may be distributed widely across $(0, 1)$ - e.g., a uniform distribution would be the common choice here.


---
## Some possible prior densities

```{r fig.height=5.3, echo=F}
x <- seq(0,1,by=0.00001)
plot(x,dbeta(x,1,1),ylim=c(0,12),xlim=c(0,1),type="l",
      col="green3",xlab=expression(theta),ylab=expression(pi(theta)))
lines(x,dbeta(x,1,10),col="red2",type="l")
lines(x,dbeta(x,2,10),col="blue2",type="l")
lines(x,dbeta(x,5,50),col="orange3",type="l")
legend("topright", legend=c("beta(1,1)","beta(1,11)","beta(2,10)","beta(5,50)"),
       col=c("green3","red2","blue2","orange3"), lwd=2, cex=1)
```


---
## Beta prior densities

- These three priors correspond to $\text{Beta}(1,1)$ (also $\text{Unif}(0,1)$), $\text{Beta}(1,10)$, $\text{Beta}(2,10)$ and $\text{Beta}(5,50)$ densities.

--

- .hlight[Beta(a,b)] is a probability density function (pdf) on (0,1),
.block[
.small[
$$\pi(\theta) = \frac{1}{B(a,b)} \theta^{a-1} (1-\theta)^{b-1},$$
]
]

  where $B(a,b)$ = beta function = normalizing constant ensuring the kernel integrates to one. Note: some texts write $\textrm{beta}(\alpha,\beta)$ instead.
  
--

- The beta(a,b) distribution has expectation $\mathbb{E}[\theta] = a/(a+b)$ and the density becomes more and more concentrated as $a + b$ = prior "sample size" increases.

--

- The variance $\mathbb{V}[\theta] = ab/[(a+b)^2(a+b+1)]$.

--

- We will look more carefully into the beta-binomial model soon but first, we will explore how this prior gets updated as data becomes available, during the online discussion session.



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!



