---
title: "STA 360/602L: Module 1.2"
subtitle: "Probability review"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```



## Outline

- Random variables

- Joint distributions

- Independence

- Exchangeability



---
## Discrete random variables

- A .hlight[random variable] is .hlight[discrete] if the set of all possible outcomes is .hlight[countable].

--

- The .hlight[probability mass function (pmf)] of a discrete random variable $Y$, $p(y)$ describes the probability associated with each possible value of $Y$.

--

- $p(y)$ has the following properties:

  1. $0 \leq p(y) \leq 1$ for all values $y \in \mathcal{Y}$.
  
  2. $\sum_{y \in \mathcal{Y}} p(y) = 1$.
  
--

- Most distributions are often charaterized by some parameter (or set/vector of parameters) $\theta$.

--

- So, to make this clear, we will often write the pmf instead as $p(y | \theta)$.


---
## Bernoulli distribution

- The .hlight[Bernoulli distribution] can be used to describe an experiment with two outcomes, such as
  + Flipping a coin (heads or tails);
  + Vote turnout (vote or not); and
  + The outcome of a basketball game (win or loss).

--

- In all cases, we can represent this as a binary random variable where the probability of "success" is $\theta$ and the probability of "failure" is $1-\theta$.

--

- We usually write this as: $Y \sim \textrm{Bernoulli}(\theta)$, where $\theta \in [0,1]$.

--

- It follows that
.block[
.small[
$$p(y|\theta) = \Pr(Y=y|\theta) = \theta^y(1-\theta)^{1-y}; \ \ \ y=0,1.$$
]
]

--

- <div class="question">
What is the mean of this distribution? What is the variance?
</div>


---
## Binomial distribution

- The .hlight[binomial distribution] describes the number of successes from $n$ independent Bernoulli trials.

--

- That is, $Y =$ number of "successes" in $n$ independent trials and $\theta$ is the probability of success per trial.

--

- We usually write this as: $Y \sim \textrm{Bin}(n,\theta)$, where $\theta \in [0,1]$.

--

- The pmf is
.block[
.small[
$$p(y|\theta) = \Pr(Y=y|\theta,n) = {n \choose y} \theta^y(1-\theta)^{n-y}; \ \ \ y=0,1,\ldots,n.$$
]
]

--

- **Example**: $Y =$ number of individuals with type I diabetes out of a sample of $n$ surveyed.

--

- Binomial likelihoods are commonly used in collecting data on proportions.

--

- <div class="question">
What is the mean of this distribution? What is the variance?
</div>


---
## Poisson distribution

- $Y \sim \textrm{Po}(\theta)$ denotes that $Y$ is a .hlight[Poisson random variable].

--

- The Poisson distribution is commonly used to model count data consisting of the number of events in a given time interval.

--

- The Poisson distribution is parameterized by $\theta$ and the pmf is given by
.block[
.small[
$$p(y|\theta) - \Pr[Y = y | \theta] = \dfrac{\theta^y e^{-\theta}}{y!}; \ \ \ \ y=0,1,2,\ldots; \ \ \ \ \theta > 0.$$
]
]

--

- Similar to binomial but with no limit on the total number of counts.

--

- <div class="question">
What is the mean of this distribution? What is the variance?
</div>


---
## General discrete distributions

- Useful to consider general discrete distributions having an arbitrary form.

--

- Suppose $Y \in \{y_1^\star,\ldots,y_k^\star\}$. Then define $\Pr(Y = y_h^\star) = \pi_h$ for each $h = 1,\ldots, k$. That is,
.block[
.small[
$$p(y|\boldsymbol{\pi}) = \Pr[Y = y| \boldsymbol{\pi}] = \prod_h \pi_h^{\mathbb{1}[Y = y_h^\star]}; \ \ y \in \{y_1^\star,\ldots,y_k^\star\}$$
]
]
  where $\boldsymbol{\pi} = (\pi_1,\ldots,\pi_k)$.

--

- $(y_1^\star,\ldots,y_k^\star)$ are "atoms" representing possible values for $Y$.

--

- For example, these may be words in a dictionary or values for education as a categorical variable. Useful for text data, categorical observations, etc.

--

- Can also write as $Y \sim \sum^k_{h=1} \pi_h \delta_{y_h^\star}$, where $\delta_{y_h^\star}$ denotes a unit mass at $y_h^\star$.

--

- Often called the .hlight[categorical distribution] or .hlight[generalized Bernoulli distribution]. Also, see the .hlight[multinomial distribution].


---
## Continuous random variables

- The .hlight[probability density function (pdf)], $p(y)$ or $f(y)$, of a continuous random variable $Y$ has slightly different properties:
  
  1. $0 \leq f(y)$ for all $y \in \mathcal{Y}$.
  
  2. $\int_{y \in \mathbb{R}} f(y) \textrm{d}y = 1$.
  
--

- The pdf for a continuous random variable is not necessarily less than 1. 

--

- Also, $f(y)$ is NOT the probability of value $y$. 

--

- However, if $f(y_1) > f(y_2)$, we say informally that $y_1$ has a "higher probability" than $y_2$.

--

- As we did in the discrete case, we will also often write the pdf instead as $f(y | \theta)$ or $p(y | \theta)$ to make the conditioning obvious.


---
## Uniform density

- The simplest example of a continuous density is the .hlight[uniform density].

--

- $Y \sim \textrm{Unif}(a,b)$ denotes density is uniform in interval $(a,b)$.

--

- The pdf is simply
.block[
.small[
$$f(y | a,b) = \dfrac{1}{b-a}; \ \ \ y  \in (a,b).$$
]
]

--

- The cdf is
.block[
.small[
$$F(y) = \Pr(Y \leq y) = \int^y_a \dfrac{1}{b-a} \textrm{d}z = \dfrac{y-a}{b-a}$$
]
]

--

- The mean (expectation) is
.block[
.small[
$$\dfrac{a+b}{2}$$
]
]

- <div class="question">
What is the variance? Also, can you prove the formula for the mean?
</div>

---
## Beta density

- The uniform density can be used as a prior for a probability if $(a,b) \subset (0,1)$.

--

- However, it is very inflexible clearly. <div class="question">
Why?
</div>

--

- An alternative for $y \in \mathcal{Y}$ is the .hlight[beta density], written as $Y \sim \textrm{Beta}(a,b)$, with
.block[
.small[
$$f(y | a,b) = \frac{1}{B(a,b)} y^{a-1} (1-y)^{b-1}; \ \ \ y  \in (0,1), \ a > 0, \ b > 0.$$
]
]

  where $B(a,b) = \dfrac{\Gamma(a)\Gamma(b)}{\Gamma(a+b)}$. $\Gamma(n) = (n-1)!$ for any positive integer $n$.
  
--

- As we have already seen, the beta density is quite flexible in characterizing a broad variety of densities on $(0,1)$.

--

- <div class="question">
Beta(1,1) is the same as Unif(0,1). Workout the pdfs to convince yourself!
</div>


---
## Gamma density

- The .hlight[gamma density] will be useful as a prior for parameters that are strictly positive.

--

- For random variables $Y \sim \textrm{Ga}(a,b)$, we have the pdf
.block[
.small[
$$f(y | a,b) = \frac{b^a}{\Gamma(a)} y^{a-1}e^{-by}; \ \ \ y  \in (0,\infty), \ a > 0, \ b > 0.$$
]
]

--

- Properties:
.block[
.small[
$$\mathbb{E}[Y] = \dfrac{a}{b}; \ \ \mathbb{V}[Y] = \dfrac{a}{b^2}.$$
]
]

--

- **Note**: there are multiple parameterizations of the gamma distribution. We will rely on this version in this course.

--

- Under this parameterization, $a$ is known as the shape parameter, while $b$ is known as the rate parameter.

--

- Under this parameterization, if $Y \sim \textrm{Ga}(1,\theta)$, then $Y \sim \textrm{Exp}(\theta)$, that is, the .hlight[exponential distribution].


---
## Continuous joint distributions

- Suppose we have two random variables $\theta = (\theta_1,\theta_2)$.

--

- Their **joint distribution function** is
.block[
.small[
$$\Pr(\theta_1 \leq a,\theta_2 \leq b) = \int^a_{-\infty} \int^b_{-\infty} f(\theta_1,\theta_2) \textrm{d}\theta_1\textrm{d}\theta_2,$$
]
]

  where $f(\theta_1,\theta_2)$ is the **joint pdf**.
  
--

- The **marginal** density of $\theta_1$ can be obtained by
.block[
.small[
$$f(\theta_1) = \int^\infty_{-\infty} f(\theta_1,\theta_2) \textrm{d}\theta_2,$$
]
]

  which is referred to as marginalizing out $\theta_2$.
  
--

- We will be doing a lot of "marginalizations", so take note!


---
## Factorizing joint densities and independence

- The joint density $f(\theta_1,\theta_2)$ can be factorized as
.block[
.small[
$$f(\theta_1,\theta_2) = f(\theta_1|\theta_2)f(\theta_2), \ \ \ \textrm{or} \ \ \ f(\theta_1,\theta_2) = f(\theta_2|\theta_1)f(\theta_1).$$
]
]

--

- For independent random variables, the joint density equals the product of the marginals:
.block[
.small[
$$f(\theta_1,\theta_2) = f(\theta_1)f(\theta_2).$$
]
]

--

- This implies that $f(\theta_2|\theta_1) = f(\theta_2)$ and $f(\theta_1|\theta_2) = f(\theta_1)$ under independence.

--

- These relationships extend automatically to $\theta = (\theta_1,\ldots,\theta_p)$. That is,
.block[
.small[
$$f(\theta_1,\ldots,\theta_p) = \prod^p_{j=1} f(\theta_j),$$
]
]

  under mutual independence of the elements of the $\theta$ vector.
  

---
## Conditional independence

- Suppose $y_i \overset{iid}{\sim} f(y_i | \theta)$ for $i = 1,\ldots,n$.

--

- Data $\{y_i\}$ are independent & identically distributed draws from distribution $f(y_i | \theta)$.

--

- The data are said to be .hlight[conditionally independent] given $\theta$ if
.block[
.small[
$$f(y_1,\ldots,y_n | \theta) = \prod^n_{i=1} f(y_i | \theta).$$
]
]

--

- $f(y_1,\ldots,y_n | \theta)$ is also the likelihood function $L(\theta | y)$ of the data.
  
--

- The .hlight[marginal likelihood] of the data is
.block[
.small[
$$L(y) = f(y_1,\ldots,y_n) = \int_\Theta f(y_1,\ldots,y_n | \theta) p(\theta)\textrm{d}\theta = \int_\Theta L(\theta | y)p(\theta)\textrm{d}\theta.$$
]
]

--

- Here, $L(y)$ can not be written as a product of densities as in $\prod\limits^n_{i=1} f(y_i)$; we lose independence when we marginalize out $\theta$.


---
## Exchangeability

- In marginalizing out $\theta$, the observations $\{y_i\}$ are not marginally independent.

--

- $\{y_i\}$ are .hlight[exchangeable] if $f(y_1,\ldots,y_n) = f(y_{\pi_1},\ldots,y_{\pi_n})$, for all permutations $\pi$ of $\{1,\ldots,n\}$.

--

- .hlight[de Finetti's Theorem]: Suppose $\{y_i\}$ are exchangeable under above definition for any $n$. Then
.block[
.small[
$$f(y_1,\ldots,y_n) = \int_\Theta \left[ \prod^n_{i=1} f(y_i| \theta) \right] p(\theta)\textrm{d}\theta.$$
]
]

  for some $\theta$, prior distribution $p(\theta)$ and sampling model $f(y_i|\theta)$.
  
--

- Simply put, de Finetti's Theorem states that exchangeable observations are conditionally independent relative to some parameter.

--

- de Finetti's Theorem is critical in providing a motivation for using parameters and for putting priors on parameters.




