---
title: "STA 360/602L: Module 2.3"
subtitle: "Marginal likelihood and posterior prediction"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Marginal likelihood

- Recall that the .hlight[marginal likelihood] is
.block[
.small[
$$L(y) = f(y_1,\ldots,y_n) = \int_\Theta f(y_1,\ldots,y_n | \theta) \pi(\theta)\textrm{d}\theta = \int_\Theta L(\theta | y)\pi(\theta)\textrm{d}\theta.$$
]
]

--

- For clarity, when dealing with a single $y$ instead of $y_1,\ldots,y_n$, we can write
.block[
.small[
$$L(y) = f(y) = \int_\Theta f(y | \theta) \pi(\theta)\textrm{d}\theta = \int_\Theta L(\theta | y)\pi(\theta)\textrm{d}\theta.$$
]
]

--

- When this is the case, for example in the case of the binomial distribution, I will often write

  + the marginal likelihood as $L(y)$ or $f(y)$, and
  
  + the sampling (conditional) likelihood as $f(y | \theta)$ or $L(\theta | y)$.


---
## Marginal likelihood

- What is the marginal likelihood for the beta-binomial?

--

- We have
.block[
.small[
$$
\begin{aligned}
L(y) & = \int_\Theta p(y|\theta)\pi(\theta)\textrm{d}\theta \\
& = \int_0^1 {n \choose y} \theta^y(1-\theta)^{n-y} \frac{1}{B(a,b)}\theta^{a-1}(1-\theta)^{b-1} d\theta\\
& = {n \choose y} \frac{B(a + y,\, b + n-y)}{B(a,b)},
\end{aligned}
$$
]
]

  by the integral definition of the Beta function.
  
--

- Marginal likelihood for the beta-Bernoulli follows directly.
  
--

- Deriving the marginal likelihood for conjugate distributions is often relatively straightforward.


---
## Prior predictive distribution

- We may care about making predictions before we even see any data.

--

- This is often useful as a way to see if the sampling distribution we have chosen is appropriate, after integrating out all unknown parameters.

--

- The .hlight[prior predictive distribution] is
.block[
.small[
$$
\begin{aligned}
p(y) &= \int_\Theta p(y,\theta)\,d\theta\\
& = \int_\Theta p(y|\theta) \cdot \pi(\theta)\,d\theta.
\end{aligned}
$$
]
]

--

- In some sense, the .hlight[prior predictive distribution] marginalizes the sampling distribution (for a single y) over the prior.

--

- When dealing with a single $y$ instead of $y_1,\ldots,y_n$, this is just the marginal likelihood of the data.
  
  

---
## Posterior predictive distribution

- We often care about making predictions for new data points, given the current pbserved data.

--

- For example, suppose $y_1,\ldots,y_n \overset{iid}{\sim} \textrm{Bernoulli}(\theta)$.

--

- We may wish to predict a new data point $y_{n+1}$.

--

- We can do so using the .hlight[posterior predictive distribution] $p(y_{n+1}|y_{1:n})$.

--

- <div class="question">
Why are we not including the parameter in the posterior predictive distribution?
</div>



---
## Posterior predictive distribution

- Recall that we have conditional independence of the $y$'s given $\theta$.

--

- So,
.block[
.small[
$$
\begin{aligned}
p(y_{n+1}|y_{1:n}) &= \int_\Theta p(y_{n+1},\theta|y_{1:n})\,d\theta\\
&= \int_\Theta p(y_{n+1}|\theta,y_{1:n}) \cdot \pi(\theta|y_{1:n})\,d\theta\\
& = \int_\Theta p(y_{n+1}|\theta) \cdot \pi(\theta|y_{1:n})\,d\theta.
\end{aligned}
$$
]
]

--

- So, in some sense, the .hlight[posterior predictive distribution] marginalizes the sampling distribution over the posterior.



---
## Posterior predictive distribution

- When we talk about the posterior predictive distribution for Bernoulli data, we are really referring to $p(y_{n+1} = 1|y_{1:n})$ and $p(y_{n+1} = 0|y_{1:n})$.

--

- Then,
.block[
.small[
$$
\begin{aligned}
p(y_{n+1}=1|y_{1:n}) &= \int_\Theta p(y_{n+1}= 1|\theta) \cdot \pi(\theta|y_{1:n})\,d\theta \\
&= ... \\
&= ...
\end{aligned}
$$
]
]

  <div class="question">
which simplifies to what? To be done on the board!
</div>
  
--

- What then is $p(y_{n+1} = 0|y_{1:n})$?

--

- Posterior predictive pmf therefore takes the form
.block[
.small[
$$p(y_{n+1}|y_{1:n}) = \dfrac{a_n^{y_{n+1}} b_n^{1-y_{n+1}}}{a_n + b_n}; \ \ \ y_{n+1}=0,1.$$
]
]

- What are $a_n$ and $b_n$?



---
## Going forward...

- From here on, we will often deal with multiple data points $y_1, \ldots, y_n$ frequently.

--

- To make that obvious, we will write the Bayes rule as one of the following
.block[
.small[
$$
\begin{split}
\pi(\theta | y_1, \ldots, y_n) & = \frac{\pi(\theta) \cdot p(y_1, \ldots, y_n|\theta)}{\int_{\Theta}\pi(\tilde{\theta}) \cdot p(y_1, \ldots, y_n| \tilde{\theta}) \textrm{d}\tilde{\theta}}\\
\\
\pi(\theta | y_1, \ldots, y_n) & = \frac{\pi(\theta) \cdot p(y_1, \ldots, y_n|\theta)}{p(y_1, \ldots, y_n)}\\
\\
\pi(\theta | y) & = \frac{\pi(\theta) \cdot L(\theta | y)}{L(y)},
\end{split}
$$
]
]

  where $y = (y_1, \ldots, y_n)$.



