---
title: "STA 360/602L: Module 3.5"
subtitle: "The normal model: joint inference for mean and variance"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```



## Joint inference for mean and variance

- We have derived the posterior for the $\mu$, conditional on $\sigma$/ $\tau$ being known. What happens when $\sigma$/ $\tau$ is unknown? We need a joint prior $\pi(\mu,\sigma^2)$ for $\mu$ and $\sigma^2$.

--

- Write the joint prior distribution for the mean and variance as the product of a conditional and a marginal distribution. That is,
.block[
.small[
$$\pi(\mu,\sigma^2) \ = \  \pi(\mu | \sigma^2)\pi(\sigma^2).$$
]
]

--

- From the previous module, we have seen that we can set the conditional prior $\pi(\mu | \sigma^2)$ to be a normal distribution.

--

- For $\pi(\sigma^2)$, we need a distribution with support on $(0,\infty)$. One such family is the gamma family, but this is NOT conjugate for the variance of a normal distribution.

--

- The gamma distribution is, however, conjugate for the precision $\tau$, and in that case, we say that $\sigma^2$ has an .hlight[inverse-gamma] distribution.



---
## Joint inference for mean and variance

- Recall that conjugacy means that for a prior $\pi(\theta)$ in a class of distributions $\mathcal{P}$ , $\pi(\theta | Y)$ is also in class $\mathcal{P}$.

--

- However, when we have multiple parameters, the dependence structure in the prior must also be preserved in the posterior, for conjugacy to hold.

--

- So, if
.block[
.small[
$$\pi(\mu,\sigma^2) \ = \  \pi(\mu | \sigma^2)\pi(\sigma^2).$$
]
]

  with $\pi(\mu | \sigma^2)$ a normal distribution, and $\pi(\sigma^2)$ an inverse-gamma distribution, we will have conjugacy if $\pi(\mu,\sigma^2 | Y)$ can also be written as
.block[
.small[
$$\pi(\mu,\sigma^2 | Y) \ = \  \pi(\mu | \sigma^2, Y)\pi(\sigma^2|Y),$$
]
]

  where $\pi(\mu | \sigma^2, Y)$ is also a normal distribution, and $\pi(\sigma^2|Y)$ is an inverse-gamma distribution, just like the prior.
  
  
---
## Inverse-gamma distribution

- As before, we will continue to work mostly in terms of the precision $\tau$.

--

- That is, we will deal with the already familiar gamma distribution, instead of the inverse-gamma distribution.

--

- However, as a quick review, if $\theta \sim \mathcal{IG}(a,b)$, then the pdf is
.block[
$$p(\theta) = \frac{b^a}{\Gamma(a)} \theta^{-(a+1)} e^{-\frac{b}{\theta} }\ \ \ \textrm{for} \ \ \ a, b > 0,$$
]

  where

  + $\mathbb{E}[\theta] = \frac{b}{a - 1}$;
  
  + $\mathbb{V}[\theta] = \frac{b^2}{(a - 1)^2(a-2)} \ \ \textrm{for} \ \ a \geq 2$.;
  
  + $\textrm{Mode}[\theta] = \frac{b}{a+1}$.



---
## Conjugate prior

- Once again, suppose $Y = (y_1,y_2,\ldots,y_n)$, where each
.block[
$$y_i \sim \mathcal{N}(\mu, \tau^{-1}).$$
]

--

- A conjugate joint prior is given by
.block[
$$
\begin{split}
\tau = \frac{1}{\sigma^2} \ & \sim \textrm{Gamma}\left(\frac{\nu_0}{2}, \frac{\nu_0\sigma_0^2}{2}\right)\\
\mu|\tau & \sim \mathcal{N}\left(\mu_0, \frac{1}{\kappa_0 \tau}\right).\\
\end{split}
$$
]

--

- This is often called a .hlight[normal-gamma] prior distribution.

--

- $\sigma_0^2$ is the prior guess for $\sigma^2$, while $\nu_0$ is often referred to as the "prior degrees of freedom", our degree of confidence in $\sigma_0^2$.

--

- We do not have conjugacy if we replace $\frac{1}{\kappa_0 \tau}$ in the normal prior with an arbitrary prior variance independent of $\tau/\sigma^2$. To do inference in that scenario, we need .hlight[Gibbs sampling] (to come soon!).



---
## Conjugate prior

- So, we have
.block[
.small[
$$\pi(\mu|\tau) = \mathcal{N}\left(\mu_0, \frac{1}{\kappa_0 \tau}\right)\propto \ \textrm{exp}\left\{-\frac{1}{2} \kappa_0 \tau (\mu - \mu_0)^2 \right\}.$$
]
]

--

- and
.block[
.small[
$$\pi(\tau) = \textrm{Ga}\left(\frac{\nu_0}{2}, \frac{\nu_0\sigma_0^2}{2}\right) \propto \tau^{\frac{\nu_0}{2}-1} \textrm{exp}\left\{-\frac{\tau\nu_0\sigma_0^2}{2}\right\}.$$
]
]

- Thus, the kernel of the normal-gamma prior distribution is
.block[
.small[
$$
\begin{split}
\Rightarrow \pi(\mu,\tau) \ & \boldsymbol{ =} \ \pi(\mu|\tau) \cdot \pi(\tau) = \mathcal{N}\left(\mu_0, \frac{1}{\kappa_0 \tau} \right) \cdot \textrm{Gamma}\left(\frac{\nu_0}{2}, \frac{\nu_0\sigma_0^2}{2}\right)\\
& \boldsymbol{ \propto} \underbrace{\textrm{exp}\left\{-\frac{1}{2} \kappa_0 \tau (\mu - \mu_0)^2 \right\}}_{\propto \ \pi(\mu|\tau)} \cdot \underbrace{\tau^{\frac{\nu_0}{2}-1} \textrm{exp}\left\{-\frac{\tau\nu_0\sigma_0^2}{2}\right\}}_{\propto \ \pi(\tau)}.\\
\end{split}
$$
]
]

--

- Take note of this form. When we derive the posterior kernel, we will try to match it to this to recognize the parameters.










---
## Posterior for the mean given variance, under normal-gamma prior

- Based on the normal-gamma prior, we need $\pi(\mu|Y,\tau)$ and $\pi(\tau|Y)$.

--

- For $\pi(\mu|Y,\tau)$, we already know from the previous module that it will be a normal distribution.

--

- However, some algebra is required to get $\pi(\tau|Y)$. 

--

- Infact, we need to write the full joint posterior and go from there, because we will need to keep some of the terms we discarded in the derivation in the last module.

--

- First, recall that the likelihood is
.block[
.large[
$$P(Y| \mu,\tau) \propto \tau^{\frac{n}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau s^2(n-1) \right\} \ \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu - \bar{y})^2 \right\}.$$
]
]


---
## Posterior derivation

.block[
.small[
$$
\begin{split}
\text{Then, } \ \ \ \ \pi(\mu,\tau | Y) \ & \boldsymbol{ \propto} \ \pi(\mu|\tau) \times \pi(\tau) \times P(Y| \mu,\tau)\\
\\
& \boldsymbol{ \propto} \underbrace{\textrm{exp}\left\{-\frac{1}{2} \kappa_0 \tau (\mu - \mu_0)^2 \right\}}_{\propto \ \pi(\mu|\sigma^2)} \times \underbrace{\tau^{\frac{\nu_0}{2}-1} \textrm{exp}\left\{-\frac{\tau\nu_0\sigma_0^2}{2}\right\}}_{\propto \ \pi(\tau)}\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \underbrace{\tau^{\frac{n}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau s^2(n-1) \right\} \ \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu - \bar{y})^2 \right\}}_{\propto \ P(Y| \mu,\tau)}\\
\\
& \boldsymbol{=}  \underbrace{\textrm{exp}\left\{-\frac{1}{2} \kappa_0 \tau (\mu - \mu_0)^2 \right\} \ \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu - \bar{y})^2 \right\}}_{\textrm{Terms involving } \mu}\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \underbrace{\tau^{\frac{\nu_0}{2}-1} \textrm{exp}\left\{-\frac{\tau\nu_0\sigma_0^2}{2}\right\} \tau^{\frac{n}{2}} \ \textrm{exp}\left\{-\frac{1}{2} \tau s^2(n-1) \right\}}_{\text{Terms involving } \tau \text{ but NOT } \mu}\\
\end{split}
$$
]
]




---
## Posterior derivation

.block[
.small[
$$
\begin{split}
\pi(\mu,\tau | Y) \ & \boldsymbol{ \propto}  \underbrace{\textrm{exp}\left\{-\frac{1}{2} \kappa_0 \tau (\mu^2 - 2\mu\mu_0 + \mu_0^2) \right\}\  \textrm{exp}\left\{-\frac{1}{2} \tau n(\mu^2 - 2\mu\bar{y} + \bar{y}^2) \right\}}_{\textrm{Terms involving } \mu}\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \underbrace{\tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}}_{\text{Terms involving } \tau \text{ but NOT } \mu}\\
& \boldsymbol{=} \ \ \underbrace{\textrm{exp}\left\{-\frac{1}{2} \left[\kappa_0 \tau (\mu^2 - 2\mu\mu_0)  + \tau n(\mu^2 - 2\mu\bar{y}) \right] \right\}}_{\textrm{Terms involving } \mu}\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \underbrace{\textrm{exp}\left\{-\frac{1}{2} \left[ \kappa_0 \tau \mu_0^2 + \tau n \bar{y}^2\right] \right\} \ \cdot  \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}}_{\text{Terms involving } \tau \text{ but NOT } \mu}\\
& \boldsymbol{=} \ \ \underbrace{\textrm{exp}\left\{-\frac{1}{2} \left[ \mu^2(n\tau + \kappa_0 \tau) - 2\mu(n\tau \bar{y} + \kappa_0 \tau\mu_0)   \right] \right\}}_{\textrm{Terms involving } \mu}\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \underbrace{\textrm{exp}\left\{-\frac{1}{2} \left[ \kappa_0 \tau \mu_0^2 + \tau n \bar{y}^2\right] \right\} \ \cdot  \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}}_{\text{Terms involving } \tau \text{ but NOT } \mu}\\
\end{split}
$$
]
]





---
## Posterior derivation

- To match the terms for the terms involving $\mu$ to the normal kernel in the prior, we need to complete the square so that we have something that looks like the $(\mu - \mu_0)^2$ term in our prior.

--

- Recall how to complete the square. Specifically, we can write
.block[
.small[
$$a\mu^2 + b\mu$$
]
]

  as
.block[
.small[
$$a(\mu + d)^2 + e,$$
]
]

  where
  + $d = \dfrac{b}{2a}$, and
  
  + $e = - \dfrac{b^2}{4a}$.




---
## Posterior derivation

- First, write out the posterior again:
.block[
.small[
$$
\begin{split}
\pi(\mu,\tau | Y) \ & \boldsymbol{=} \ \ \underbrace{\textrm{exp}\left\{-\frac{1}{2} \left[ (n\tau + \kappa_0 \tau)\mu^2 - 2\mu(n\tau \bar{y} + \kappa_0 \tau\mu_0)   \right] \right\}}_{\textrm{Terms involving } \mu}\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \underbrace{\textrm{exp}\left\{-\frac{1}{2} \left[ \kappa_0 \tau \mu_0^2 + \tau n \bar{y}^2\right] \right\} \ \cdot  \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}}_{\text{Terms involving } \tau \text{ but NOT } \mu}\\
\end{split}
$$
]
]


- Set $a^\star = (n\tau + \kappa_0 \tau)$ and $b^\star = (n\tau \bar{y} + \kappa_0 \tau\mu_0)$, then complete the square for the first part.
.block[
.small[
$$
\begin{split}
\Rightarrow \ \pi(\mu,\tau | Y) \ & \boldsymbol{ \propto} \ \ \underbrace{\textrm{exp}\left\{-\frac{1}{2} \left[ a^\star \mu^2 - 2 b^\star \mu   \right] \right\}}_{\textrm{Terms involving } \mu}\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \underbrace{\textrm{exp}\left\{-\frac{1}{2} \left[ \kappa_0 \tau \mu_0^2 + \tau n \bar{y}^2\right] \right\} \ \cdot  \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}}_{\text{Terms involving } \tau \text{ but NOT } \mu}\\
\end{split}
$$
]
]



---
## Posterior derivation

.block[
.small[
$$
\begin{split}
\Rightarrow \ \pi(\mu,\tau | Y) \ & \boldsymbol{ \propto} \ \ \textrm{exp}\left\{-\frac{1}{2} a^\star \left[ \mu- \frac{b^\star}{a^\star} \right]^2 + \frac{(b^\star)^2}{2a^\star} \right\} \ \cdot \ \textrm{exp}\left\{-\frac{1}{2} \left[ \kappa_0 \tau \mu_0^2 + \tau n \bar{y}^2\right] \right\}\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}\\
\\
& \boldsymbol{=} \ \ \textrm{exp}\left\{-\frac{1}{2} a^\star \left[ \mu- \frac{b^\star}{a^\star} \right]^2 \right\} \ \underbrace{\textrm{exp}\left\{-\frac{1}{2} \left[ \kappa_0 \tau \mu_0^2 + \tau n \bar{y}^2 - \frac{(b^\star)^2}{a^\star}\right] \right\}}_{\textrm{Next, substitute the values for } a^\star \textrm{ and } b^\star \textrm{ back}}\\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}\\
\\
& \boldsymbol{=} \ \ \textrm{exp}\left\{-\frac{1}{2} a^\star \left[ \mu- \frac{b^\star}{a^\star} \right]^2 \right\} \ \textrm{exp} \underbrace{\left\{-\frac{1}{2} \left[ \kappa_0 \tau \mu_0^2 + \tau n \bar{y}^2 - \frac{(n\tau \bar{y} + \kappa_0 \tau\mu_0)^2}{(n\tau + \kappa_0 \tau)}\right] \right\}}_{\textrm{Next, expand terms and recombine}} \\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}\\
\end{split}
$$ 
]
]


---
## Posterior derivation

.block[
.small[
$$
\begin{split}
\Rightarrow \ \pi(\mu,\tau | Y) \ & \boldsymbol{ \propto} \ \ \textrm{exp}\left\{-\frac{1}{2} a^\star \left[ \mu- \frac{b^\star}{a^\star} \right]^2 \right\} \ \textrm{exp}\left\{-\frac{1}{2} \left[\frac{n\kappa_0\tau^2 (\mu_0^2- 2\mu_0\bar{y} + \bar{y}^2)}{\tau(\kappa_0 + n)}\right] \right\}\\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}\\
\\
& \boldsymbol{=} \ \ \textrm{exp}\left\{-\frac{1}{2} a^\star \left[ \mu- \frac{b^\star}{a^\star} \right]^2 \right\} \ \textrm{exp}\left\{-\frac{\tau}{2} \left[\frac{n\kappa_0 (\bar{y} - \mu_0)^2}{(\kappa_0 + n)}\right] \right\}\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\}\\
 \\
& \boldsymbol{=} \ \ \textrm{exp} \underbrace{\left\{-\frac{1}{2} a^\star \left[ \mu- \frac{b^\star}{a^\star} \right]^2 \right\}}_{\textrm{Substitute the values for } a^\star \textrm{ and } b^\star \textrm{ back}} \\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau \left[\nu_0\sigma_0^2 + s^2(n-1) \right] }{2}\right\} \textrm{exp}\left\{-\frac{\tau}{2} \left[\frac{n\kappa_0 (\bar{y} - \mu_0)^2}{(\kappa_0 + n)}\right] \right\}\\
\end{split}
$$ 
]
]




---
## Posterior derivation

.block[
.small[
$$
\begin{split}
\Rightarrow \ \pi(\mu,\tau | Y) \ & \boldsymbol{ \propto} \ \ \underbrace{\textrm{exp}\left\{-\frac{1}{2} (n\tau + \kappa_0 \tau) \left[ \mu- \frac{(n\tau \bar{y} + \kappa_0 \tau\mu_0)}{(n\tau + \kappa_0 \tau)} \right]^2 \right\}}_{\textrm{Normal Kernel}}\\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \underbrace{\tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau}{2} \left[\nu_0\sigma_0^2 + s^2(n-1) + \frac{n\kappa_0}{(\kappa_0 + n)} (\bar{y} - \mu_0)^2 \right] \right\}}_{\textrm{Gamma Kernel}}\\
\\
& \boldsymbol{=} \ \ \underbrace{\textrm{exp}\left\{-\frac{1}{2} \tau(\kappa_0 + n) \left[ \mu - \frac{(\kappa_0 \mu_0 + n \bar{y})}{(\kappa_0 + n)} \right]^2 \right\}}_{\textrm{Normal Kernel}}\\\
& \ \ \ \ \ \ \ \ \ \ \ \ 
\times \underbrace{\tau^{\frac{\nu_0 + n}{2}-1} \textrm{exp}\left\{-\frac{\tau}{2} \left[\nu_0\sigma_0^2 + s^2(n-1) + \frac{n\kappa_0}{(\kappa_0 + n)} (\bar{y} - \mu_0)^2 \right] \right\}}_{\textrm{Gamma Kernel}}\\
\end{split}
$$
]
]



---
## Posterior derivation

.block[
.small[
$$
\begin{split}
\Rightarrow \ \pi(\mu,\tau | Y) \ & \boldsymbol{=} \ \ \mathcal{N}\left(\mu_n, \frac{1}{\kappa_n\tau} \right) \times \textrm{Gamma}\left(\frac{\nu_n}{2}, \frac{\nu_n \sigma_n^2}{2}\right) = \pi(\mu|Y, \tau) \pi(\tau|Y),
\end{split}
$$
]
]

   where
.block[
.small[
$$
\begin{split}
\kappa_n & = \kappa_0 + n\\
\mu_n & = \frac{\kappa_0 \mu_0 + n\bar{y}}{\kappa_n} = \frac{\kappa_0}{\kappa_n} \mu_0 + \frac{n}{\kappa_n} \bar{y}\\
\nu_n & = \nu_0 + n\\
\sigma_n^2 & = \frac{1}{\nu_n}\left[\nu_0\sigma_0^2 + s^2(n-1) + \frac{n\kappa_0}{\kappa_n} (\bar{y} - \mu_0)^2 \right] = \frac{1}{\nu_n}\left[\nu_0\sigma_0^2 + \sum_{i=1}^n (y_i-\bar{y})^2 + \frac{n\kappa_0}{\kappa_n} (\bar{y} - \mu_0)^2 \right]\\
\end{split}
$$
]
]

--

- Turns out that the marginal posterior of $\mu$, that is, $\pi(\mu|Y) = \int_0^\infty \pi(\mu, \tau|Y) d\tau$ is a **t-distribution**. 

--

- You can derive that distribution if you are interested, we won't spend time on it in class. We will be able to sample from it through Monte Carlo anyway.


---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




