---
title: "STA 360/602L: Module 3.2"
subtitle: "Rejection sampling; Importance sampling"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Rejection sampling

- .hlight[Rejection sampling] and .hlight[Importance sampling] are one of the first steps into Monte Carlo analysis, in which simulated values from one distribution are used to explore another.

--

- Simulating from the "wrong distribution" can be incredibly useful as we will see in this module and also later in the course.

--

- Both are not used very often, but are still of practical interest in
  + fairly small problems, in terms of dimension, 
  
--
  + in which the density of the distribution of interest can be easily evaluated, but when it is difficult to sample from directly, and
  
--
  + when it is relatively easy to identify and simulate from distributions that approximate the distribution of interest.
  
--

- Importance sampling and Rejection sampling use the same ideas, but the latter leads to exact corrections and so exact samples from the distribution of interest.


---
## Rejection sampling

- Setup:
  + $p(\theta)$ is some density we are interested in sampling from;
  
--
  + $p(\theta)$ is tough to sample from but we are able to evaluate $p(\theta)$ as a function at any point; and
  
--
  + $g(\theta)$ is some .hlight[proposal distribution] or .hlight[importance sampling distribution] that is easier to sample from.

--

- Two key requirements:
  + $g(\theta)$ is easy to sample from; and
  
--
  + $g(\theta)$ is easy to evaluate at any point as is the case for $p(\theta)$.
  
--

- Usually, the context is one in which $g(\theta)$ has been derived as an analytic approximation to $p(\theta)$; and the closer the approximation, the more accurate the resulting Monte Carlo analysis will be.



---
## Rejection sampling

- Procedure:

  1. Define $w(\theta) = p(\theta)/g(\theta)$.
  
--
  2. Assume that $w(\theta) = p(\theta)/g(\theta) < M$ for some constant M. If $g(\theta)$ represents a good approximation to $p(\theta)$, then M should not be too far from 1.

--
  3. Generate a _candidate_ value $\theta \sim g(\theta)$ and **accept** with probability $w(\theta)/M$: if accepted, $\theta$ is a draw from $p(\theta)$; otherwise **reject** and try again.  
  
--
      Equivalently, generate $u \sim U(0,1)$ independently of $\theta$. Then **accept** $\theta$ as a draw from $p(\theta)$ if, and only if, $u < w(\theta)/M$.

--

- For those interested, the proof that all accepted $\theta$ values are indeed from $p(\theta)$ is on the next slide. We will not spend time on it.

--

- Clearly, we need $M$ for this to work. However, in the case of truncated densities, we actually have $M$.



---
## Proof for simple accept/reject

- We need to show that all accepted $\theta$ values are indeed from $p(\theta)$. Equivalently, show that $f(\theta | u < w(\theta)/M) = p(\theta)$.

- By Bayes' theorem,
.block[
.small[
$$f(\theta | u < w(\theta)/M) = \dfrac{\Pr(\theta \textrm{ and } u < w(\theta)/M)}{\Pr(u < w(\theta)/M)} = \dfrac{\Pr(u < w(\theta)/M \ | \ \theta) g(\theta)}{\Pr(u < w(\theta)/M)}.$$
]
]

- But,
  + $\Pr(u < w(\theta)/M \ | \ \theta) = w(\theta)/M$ since $u \sim U(0,1)$, and
  
  + 
.block[
.small[
$$
\begin{split}
\Pr(u < w(\theta)/M) & = \int \Pr(u < w(\theta)/M \ | \ \theta) g(\theta) d\theta\\
& = \int w(\theta)/M g(\theta) d\theta = 1/M \int w(\theta) g(\theta) d\theta = 1/M \int p(\theta) d\theta = 1/M.
\end{split}
$$
]
]


- Therefore,
.block[
.small[
$$f(\theta | u < w(\theta)/M) = \dfrac{\Pr(u < w(\theta)/M \ | \ \theta) g(\theta)}{\Pr(u < w(\theta)/M)} = \dfrac{w(\theta)/M g(\theta)}{1/M} = w(\theta) g(\theta) = p(\theta).$$
]
]



---
## Rejection sampling for truncated densities

- The inverse CDF method works well for truncated densities but what happens when we can not or prefer not to write down the truncated CDF?

--

- Suppose we want to sample from $f_{[a,b]}(\theta)$, that is, a known pdf $f(\theta)$ truncated to $[a,b]$.

  + Recall that $f_{[a,b]}(\theta) \propto f(\theta)\mathbb{1}[\theta \in [a,b]]$. Using the notation for rejection sampling, $p(\theta) = f_{[a,b]}(\theta)$ and $g(\theta) = f(\theta)$.
  
--

  + Set $1/M = \int^b_a f(\theta^\star)\textrm{d}\theta^\star$, so that $M$ is the normalizing constant of the truncated density.
  
--

  + Then, $w(\theta) = p(\theta)/g(\theta) = M \mathbb{1}[\theta \in [a,b]] \leq M$ as required.
  
  
  
---
## Rejection sampling for truncated densities

- We can then use the procedure on page 5 to generate the required samples. 

--

- Specifically,

--
  + For each $i=1,\ldots,m$, generate $\theta_i \sim f$. If $\theta_i \in [a,b]$, accept $\theta_i$, otherwise **reject** and try again. 
  
--
  + Easy to show that this is equivalent to accepting each $\theta_i$ with probability $w(\theta)/M$.





---
## Example

```{r}
#Simple code for using rejection sampling to generate m samples
#from the Beta[10,10] density truncated to (0.35,0.6).
set.seed(12345)
#NOTE: there are more efficient ways to write this code!

#set sample size and reate vector to store sample
m <- 10000; THETA <- rep(0,m)
#keep track of rejects
TotalRejects <- 0; Rejections <- NULL
#now the 'for loop'
for(i in 1:m){
  t <- 0
  while(t < 1){
    theta <- rbeta(1,10,10)
    if(theta > 0.35 & theta < 0.6){
      THETA[i] <- theta
      t <- 1
    } else {
    TotalRejects <- TotalRejects + 1
    Rejections <- rbind(Rejections,theta)
  }
}
}
#Overall acceptance rate:
1 - TotalRejects/(m+TotalRejects)
```


---
## Example

How does our sample compare to the true truncated density? $m = 100$
```{r fig.height=4.5, echo=F}
set.seed(12345)
m <- 100; THETA <- rep(0,m)
TotalRejects <- 0; Rejections <- NULL
for(i in 1:m){
  t <- 0
  while(t < 1){
    theta <- rbeta(1,10,10)
    if(theta > 0.35 & theta < 0.6){
      THETA[i] <- theta
      t <- 1
    } else {
    TotalRejects <- TotalRejects + 1
    Rejections <- rbind(Rejections,theta)
  }
}
}

p <- seq(0,1,length=10000)
f1 <- dbeta(p,10,10)*as.numeric(p>0.35 & p<0.6)/ (pbeta(0.6,10,10) - pbeta(0.3,10,10) )
plot(p,f1,type='l',col='green4',xlim=c(0,1),ylab='Density', xlab=expression(theta),
     ylim=c(0,6),lty=1.5)
lines(density(THETA),type='l',col='blue4',lty=1.5)
points(c(Rejections),rep(0,TotalRejects),type='p',col="red2",pch=4)
labels <- c("True Density", "Accepted Samples","Rejects")
legend("topright", labels, lwd=c(2,2,2), lty=c(1.5,1.5,NA),pch=c(NA,NA,4),
       col=c('green4','blue4',"red2"))
```



---
## Example

How does our sample compare to the true truncated density? $m = 1000$
```{r fig.height=4.5, echo=F}
set.seed(12345)
m <- 1000; THETA <- rep(0,m)
TotalRejects <- 0; Rejections <- NULL
for(i in 1:m){
  t <- 0
  while(t < 1){
    theta <- rbeta(1,10,10)
    if(theta > 0.35 & theta < 0.6){
      THETA[i] <- theta
      t <- 1
    } else {
    TotalRejects <- TotalRejects + 1
    Rejections <- rbind(Rejections,theta)
  }
}
}

p <- seq(0,1,length=10000)
f1 <- dbeta(p,10,10)*as.numeric(p>0.35 & p<0.6)/ (pbeta(0.6,10,10) - pbeta(0.3,10,10) )
plot(p,f1,type='l',col='green4',xlim=c(0,1),ylab='Density', xlab=expression(theta),
     ylim=c(0,6),lty=1.5)
lines(density(THETA),type='l',col='blue4',lty=1.5)
points(c(Rejections),rep(0,TotalRejects),type='p',col="red2",pch=4)
labels <- c("True Density", "Accepted Samples","Rejects")
legend("topright", labels, lwd=c(2,2,2), lty=c(1.5,1.5,NA),pch=c(NA,NA,4),
       col=c('green4','blue4',"red2"))
```




---
## Example

How does our sample compare to the true truncated density? $m = 10000$
```{r fig.height=4.5, echo=F, cache=T}
set.seed(12345)
m <- 10000; THETA <- rep(0,m)
TotalRejects <- 0; Rejections <- NULL
for(i in 1:m){
  t <- 0
  while(t < 1){
    theta <- rbeta(1,10,10)
    if(theta > 0.35 & theta < 0.6){
      THETA[i] <- theta
      t <- 1
    } else {
    TotalRejects <- TotalRejects + 1
    Rejections <- rbind(Rejections,theta)
  }
}
}

p <- seq(0,1,length=10000)
f1 <- dbeta(p,10,10)*as.numeric(p>0.35 & p<0.6)/ (pbeta(0.6,10,10) - pbeta(0.3,10,10) )
plot(p,f1,type='l',col='green4',xlim=c(0,1),ylab='Density', xlab=expression(theta),
     ylim=c(0,6),lty=1.5)
lines(density(THETA),type='l',col='blue4',lty=1.5)
points(c(Rejections),rep(0,TotalRejects),type='p',col="red2",pch=4)
labels <- c("True Density", "Accepted Samples","Rejects")
legend("topright", labels, lwd=c(2,2,2), lty=c(1.5,1.5,NA),pch=c(NA,NA,4),
       col=c('green4','blue4',"red2"))
```



---
## Example

How does our sample compare to the true truncated density? $m = 100000$
```{r fig.height=4.5, echo=F, cache=T}
set.seed(12345)
m <- 100000; THETA <- rep(0,m)
TotalRejects <- 0; Rejections <- NULL
for(i in 1:m){
  t <- 0
  while(t < 1){
    theta <- rbeta(1,10,10)
    if(theta > 0.35 & theta < 0.6){
      THETA[i] <- theta
      t <- 1
    } else {
    TotalRejects <- TotalRejects + 1
    Rejections <- rbind(Rejections,theta)
  }
}
}

p <- seq(0,1,length=10000)
f1 <- dbeta(p,10,10)*as.numeric(p>0.35 & p<0.6)/ (pbeta(0.6,10,10) - pbeta(0.3,10,10) )
plot(p,f1,type='l',col='green4',xlim=c(0,1),ylab='Density', xlab=expression(theta),
     ylim=c(0,6),lty=1.5)
lines(density(THETA),type='l',col='blue4',lty=1.5)
points(c(Rejections),rep(0,TotalRejects),type='p',col="red2",pch=4)
labels <- c("True Density", "Accepted Samples","Rejects")
legend("topright", labels, lwd=c(2,2,2), lty=c(1.5,1.5,NA),pch=c(NA,NA,4),
       col=c('green4','blue4',"red2"))
```



---
## Comments

- Clearly less efficient than the inverse CDF method, which we already know how to use for this particular problem.

--

- When you can write down the truncated CDF, use the inverse CDF method instead.

--

- When you cannot, rejection sampling can be a possible alternative, as are many more sampling methods which we will not cover in this course.

--

- Anyway, generally, rejection sampling can still be very useful.

--

- Importance sampling is another related sampling method but we will not spend time on it. If you are interested, take a look at the next few slides. If not, feel free to skip.



---
class: center, middle

# Optional content from here on...


---
## Importance sampling

- Interest lies in expectations of the form (instead of the actual samples)
.block[
$$H = \int h(\theta) p(\theta) d\theta,$$
]

- Write
.block[
$$H = \int h(\theta) w(\theta) g(\theta) d\theta \ \ \ \textrm{with} \ \ \ w(\theta) = p(\theta)/g(\theta)$$
]

  that is, $\mathbb{E}[h(\theta)]$ under $p(\theta)$ is just $\mathbb{E}[h(\theta) w(\theta)]$ under $g(\theta)$.
  
- Using direct Monte Carlo integration
.block[
$$\bar{h} = \dfrac{1}{m} \sum_{i=1}^m w(\theta_i) h(\theta_i).$$
]

  where $\theta_1, \ldots, \theta_m \overset{ind}{\sim} g(\theta)$. We are sampling from the "wrong" distribution.



---
## Importance sampling

- The measure of "how wrong" we are at each simulated $\theta_m$ value is the .hlight[importance weight]
.block[
$$w(\theta_i) = p(\theta_i)/g(\theta_i).$$
]

--

- These ratios weight the sample estimates $h(\theta_i)$ to "correct" for the fact that we sampled the wrong distribution.
  
- See [Lopes & Gamerman (Ch 3.4)](https://www.amazon.com/Markov-Chain-Monte-Carlo-Statistical/dp/1584885874) and [Robert and Casella (Ch. 3.3)](https://www.amazon.com/Monte-Statistical-Methods-Springer-Statistics/dp/1441919392) for discussions of convergence and optimality.

--

- Clearly, the closer $g$ is to $p$, the better the results, just as we had with rejection sampling.



---
## Importance sampling

- Key considerations:
  + MC estimate $\bar{h}$ has the expectation $H$; and is generally almost surely convergent to $H$ (under certain conditions of course but we will not dive into those).
  
--
  + $\mathbb{V}[\bar{h}]$ is often going to be finite in cases in which, generally, $w(\theta) = p(\theta)/g(\theta)$ is bounded and decays rapidly in the tails of $p(\theta)$.
  
--
  + Thus, superior MC approximations, are achieved for choices of $g(\theta)$ whose tails dominate those of the target $p(\theta)$.
  
--
  + That is, importance sampling distributions should be chosen to have tails at least as fat as the target (think normal distribution vs t-distribution).
  
--
  + Obviously require the support of $g(\theta)$ to be the same as, or contain, that of $p(\theta)$.
  
--

- These also clearly apply to rejection sampling too.
  
  
  
---
## Importance sampling

- Problems in which $w(\theta) = p(\theta)/g(\theta)$ can be computed are actually rare.

--

- As you will see when we move away from conjugate distributions, we usually only know $p(\theta)$ up to a normalizing constant.

--

- When this is the case, simply "re-normalize" the importance weights, so that
.block[
$$\bar{h} = \dfrac{1}{m} \sum_{i=1}^m w_i h(\theta_i) \ \ \ \textrm{where} \ \ \ w_i = \dfrac{w(\theta_i)}{\sum^m_{i=1}w(\theta_i)}.$$
]

--

- Generally, in importance sampling, weights that are close to uniform are desirable, and very unevenly distributed weights are not.



---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




