---
title: "STA 360/602L: Module 4.5"
subtitle: "Missing data and imputation I"
author: "Dr. Olanrewaju Michael Akande"
date: " "
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```


## Introduction to missing data

- Missing data/nonresponse is fairly common in real data. For example,
  + Failure to respond to survey question
  + Subject misses some clinic visits out of all possible
  + Only subset of subjects asked certain questions
  
--

- Recall that our posterior computation usually depends on the data through $\mathcal{p}(Y| \theta)$, which cannot be
computed (at least directly) when some of the $y_i$ values are missing.

--

- The most common software packages often throw away all subjects with incomplete data (can lead to bias and precision loss).

--

- Some individuals impute missing values with a mean or some other fixed value (ignores uncertainty).

--

- As you will see, imputing missing data is actually quite natural in the Bayesian context.



<!-- --- -->
<!-- ## Main types of nonresponse -->

<!-- - .hlight[Unit nonresponse]: the individual has no values recorded for any of the variables. For example, when participants do not complete a survey questionnaire at all. -->

<!-- - .hlight[Item nonresponse]: the individual has values recorded for at least one variable, but not all variables. -->

<!-- <table> -->
<!--   <caption>Unit nonresponse vs item nonresponse</caption> -->
<!--   <tr> -->
<!--     <th> </th> -->
<!--     <th height="30px" colspan="3">Variables</th> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <th>  </th> -->
<!--     <td height="30px" style="text-align:center" width="100px"> Y<sub>1</sub> </td> -->
<!--     <td style="text-align:center" width="100px"> Y<sub>2</sub> </td> -->
<!--     <td style="text-align:center" width="100px"> Y<sub>3</sub> </td> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <td height="30px" style="text-align:left"> Complete cases </td> -->
<!--     <td style="text-align:center"> &#10004 </td> -->
<!--     <td style="text-align:center"> &#10004 </td> -->
<!--     <td style="text-align:center"> &#10004 </td> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <td rowspan="3"> Item nonresponse </td> -->
<!--     <td rowspan="3" style="text-align:center"> &#10004 </td> -->
<!--     <td height="30px" style="text-align:center"> &#10004 </td> -->
<!--     <td style="text-align:center"> &#10067 </td> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <td height="30px" style="text-align:center"> &#10067 </td> -->
<!--     <td style="text-align:center"> &#10067 </td> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <td height="30px" style="text-align:center"> &#10067 </td> -->
<!--     <td style="text-align:center"> &#10004 </td> -->
<!--   </tr> -->
<!--   <tr> -->
<!--     <td height="30px" style="text-align:left"> Unit nonresponse </td> -->
<!--     <td style="text-align:center"> &#10067 </td> -->
<!--     <td style="text-align:center"> &#10067 </td> -->
<!--     <td style="text-align:center"> &#10067 </td> -->
<!--   </tr> -->
<!-- </table> -->



---
## Missing data mechanisms

- Data are said to be .hlight[missing completely at random (MCAR)] if the reason for missingness does not depend on the values of the observed data or missing data.

--

- For example, suppose
  - you handed out a double-sided survey questionnaire of 20 questions to a sample of participants;
  - questions 1-15 were on the first page but questions 16-20 were at the back; and
  - some of the participants did not respond to questions 16-20.
 
--
 
- Then, the values for questions 16-20 for those people who did not respond would be .hlight[MCAR] if they simply did not realize the pages were double-sided; they had no reason to ignore those questions.
 
--
 
- **This is rarely plausible in practice!**


---
## Missing data mechanisms

- Data are said to be .hlight[missing at random (MAR)] if, conditional on the values of the observed data, the reason for missingness does not depend on the missing data.

--

- Using our previous example, suppose
  - questions 1-15 include demographic information such as age and education;
  - questions 16-20 include income related questions; and
  - once again, some participants did not respond to questions 16-20.

--
  
- Then, the values for questions 16-20 for those people who did not respond would be .hlight[MAR] if younger people are more likely not to respond to those income related questions than old people, where age is observed for all participants.
  
--

- **This is the most commonly assumed mechanism in practice!**


---
## Missing data mechanisms

- Data are said to be .hlight[missing not at random (MNAR or NMAR)] if the reason for missingness depends on the actual values of the missing (unobserved) data.

--

- Continuing with our previous example, suppose again that
  - questions 1-15 include demographic information such as age and education;
  - questions 16-20 include income related questions; and
  - once again, some of the participants did not respond to questions 16-20.

--
  
- Then, the values for questions 16-20 for those people who did not respond would be .hlight[MNAR] if people who earn more money are less likely to respond to those income related questions than old people.

--
  
- **This is usually the case in real data, but analysis can be complex!**


---
## Mathematical formulation

- Consider the multivariate data scenario with $\boldsymbol{Y}_i = (\boldsymbol{Y}_1,\ldots,\boldsymbol{Y}_n)^T$, where $\boldsymbol{Y}_i = (Y_{i1},\ldots,Y_{ip})^T$, for $i = 1,\ldots, n$.

--

- For now, we will assume the multivariate normal model as the sampling model, so that each $\boldsymbol{Y}_i = (Y_{i1},\ldots,Y_{ip})^T \sim \mathcal{N}_p(\boldsymbol{\theta}, \Sigma)$.

--
	
- Suppose now that $\boldsymbol{Y}$ contains missing values.

--

- We can separate $\boldsymbol{Y}$ into the observed and missing parts, that is, $\boldsymbol{Y} = (\boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis})$.

--

- Then for each individual, $\boldsymbol{Y}_i = (\boldsymbol{Y}_{i,obs},\boldsymbol{Y}_{i,mis})$.



---
## Mathematical Formulation

- Let
  + $j$ index variables (where $i$ already indexes individuals),
  + $r_{ij} = 1$ when $y_{ij}$ is missing,
  + $r_{ij} = 0$ when $y_{ij}$ is observed.

--

- Here, $r_{ij}$ is known as the missingness indicator of variable $j$ for person $i$. 

--

- Also, let 
  + $\boldsymbol{R}_i = (r_{i1},\ldots,r_{ip})^T$ be the vector of missing indicators for person $i$.
  + $\boldsymbol{R} = (\boldsymbol{R}_1,\ldots,\boldsymbol{R}_n)$ be the matrix of missing indicators for everyone.
  + $\boldsymbol{\psi}$ be the set of parameters associated with $\boldsymbol{R}$.

--

- Assume $\boldsymbol{\psi}$ and $(\boldsymbol{\theta}, \Sigma)$ are distinct.


---
## Mathematical Formulation

- MCAR:
.block[
$$p(\boldsymbol{R} | \boldsymbol{Y},\boldsymbol{\theta}, \Sigma, \boldsymbol{\psi}) = p(\boldsymbol{R} | \boldsymbol{\psi})$$
]

--

- MAR:
.block[
$$p(\boldsymbol{R} | \boldsymbol{Y},\boldsymbol{\theta}, \Sigma, \boldsymbol{\psi}) = p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{\psi})$$
]

--

- MNAR:
.block[
$$p(\boldsymbol{R} | \boldsymbol{Y},\boldsymbol{\theta}, \Sigma, \boldsymbol{\psi}) = p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis},\boldsymbol{\psi})$$
]



---
## Implications for likelihood function

- Each type of mechanism has a different implication on the likelihood of the observed data $\boldsymbol{Y}_{obs}$, and the missing data indicator $\boldsymbol{R}$.

--

- Without missingness in $\boldsymbol{Y}$, the likelihood of the observed data is
.block[
$$p(\boldsymbol{Y}_{obs} | \boldsymbol{\theta}, \Sigma)$$
]

--

- With missingness in $\boldsymbol{Y}$, the likelihood of the observed data is instead
.block[
$$
\begin{split}
p(\boldsymbol{Y}_{obs}, \boldsymbol{R} |\boldsymbol{\theta}, \Sigma, \boldsymbol{\psi}) & = \int p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis},\boldsymbol{\psi}) \cdot p(\boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis} | \boldsymbol{\theta}, \Sigma) \textrm{d}\boldsymbol{Y}_{mis} \\
\end{split}
$$
]

 <!-- = \prod^n_{i=1}  p(\boldsymbol{Y}_{i,obs} | \boldsymbol{\theta}, \Sigma) -->

<!-- & = \int p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis},\boldsymbol{\psi}) \prod^n_{i=1}  p(\boldsymbol{Y}_{i,obs}, \boldsymbol{Y}_{i,mis} | \boldsymbol{\theta}, \Sigma) \textrm{d}\boldsymbol{Y}_{mis} \\ -->

--

- Since we do not actually observe $\boldsymbol{Y}_{mis}$, we would like to be able to integrate it out so we don't have to deal with it.

--

- That is, we would like to infer $(\boldsymbol{\theta}, \Sigma)$ (and sometimes, $\boldsymbol{\psi}$) using only the observed data.



---
## Likelihood function: MCAR

- For MCAR, we have:
.block[
$$
\begin{split}
p(\boldsymbol{Y}_{obs}, \boldsymbol{R} |\boldsymbol{\theta}, \Sigma, \boldsymbol{\psi}) & = \int p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis},\boldsymbol{\psi}) \cdot p(\boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis} | \boldsymbol{\theta}, \Sigma) \textrm{d}\boldsymbol{Y}_{mis} \\
& = \int p(\boldsymbol{R} | \boldsymbol{\psi}) \cdot p(\boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis} | \boldsymbol{\theta}, \Sigma) \textrm{d}\boldsymbol{Y}_{mis} \\
& = p(\boldsymbol{R} | \boldsymbol{\psi}) \cdot \int p(\boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis} | \boldsymbol{\theta}, \Sigma) \textrm{d}\boldsymbol{Y}_{mis} \\
& = p(\boldsymbol{R} | \boldsymbol{\psi}) \cdot p(\boldsymbol{Y}_{obs} | \boldsymbol{\theta}, \Sigma). \\
\end{split}
$$
]

--

- For inference on $(\boldsymbol{\theta}, \Sigma)$, we can simply focus on $p(\boldsymbol{Y}_{obs} | \boldsymbol{\theta}, \Sigma)$ in the likelihood function, since $(\boldsymbol{R} | \boldsymbol{\psi})$ does not include any $\boldsymbol{Y}$.



---
## Likelihood function: MAR

- For MAR, we have:
.block[
$$
\begin{split}
p(\boldsymbol{Y}_{obs}, \boldsymbol{R} |\boldsymbol{\theta}, \Sigma, \boldsymbol{\psi}) & = \int p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis},\boldsymbol{\psi}) \cdot p(\boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis} | \boldsymbol{\theta}, \Sigma) \textrm{d}\boldsymbol{Y}_{mis} \\
& = \int p(\boldsymbol{R} | \boldsymbol{Y}_{obs}, \boldsymbol{\psi}) \cdot p(\boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis} | \boldsymbol{\theta}, \Sigma) \textrm{d}\boldsymbol{Y}_{mis} \\
& = p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{\psi}) \cdot \int p(\boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis} | \boldsymbol{\theta}, \Sigma) \textrm{d}\boldsymbol{Y}_{mis} \\
& = p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{\psi}) \cdot p(\boldsymbol{Y}_{obs} | \boldsymbol{\theta}, \Sigma). \\
\end{split}
$$
]

--

- For inference on $(\boldsymbol{\theta}, \Sigma)$, we can once again focus on $p(\boldsymbol{Y}_{obs} | \boldsymbol{\theta}, \Sigma)$ in the likelihood function,  although there can be some bias if we do not account for $p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{\psi})$, since it contains observed data.

--

- Also, if we want to infer the missingness mechanism through $\boldsymbol{\psi}$, we would need to deal with $p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{\psi})$ anyway.


---
## Likelihood function: MNAR

- For MNAR, we have:
.block[
$$
\begin{split}
p(\boldsymbol{Y}_{obs}, \boldsymbol{R} |\boldsymbol{\theta}, \Sigma, \boldsymbol{\psi}) & = \int p(\boldsymbol{R} | \boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis},\boldsymbol{\psi}) \cdot p(\boldsymbol{Y}_{obs},\boldsymbol{Y}_{mis} | \boldsymbol{\theta}, \Sigma) \textrm{d}\boldsymbol{Y}_{mis} \\
\end{split}
$$
]

--

- The likelihood under MNAR cannot simplify any further.
  
--

- In this case, we cannot ignore the missing data when making inferences about $(\boldsymbol{\theta}, \Sigma)$.
  
--

- We must include the model for $\boldsymbol{R}$ and also infer the missing data $\boldsymbol{Y}_{mis}$.
  
  

---
## How to tell in practice?

- So how can we tell the type of mechanism we are dealing with? 

--

- In general, we don't know!!!

--

- Rare that data are MCAR (unless planned beforehand); more likely that data are MNAR.

--

- **Compromise**: assume data are MAR if we include enough variables in model for the missing data indicator $\boldsymbol{R}$.

--

- Whenever we talk about missing data in this course, we will do so in the context of MCAR and MAR.
  

---

class: center, middle

# What's next? 

### Move on to the readings for the next module!




