---
title: "The multinomial model"
author: "Dr. Olanrewaju Michael Akande"
date: "Feb 12, 2020"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/class_logo.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
knitr::opts_chunk$set(fig.height = 2.65, dpi = 300,fig.align='center',fig.show='hold',size='footnotesize', small.mar=TRUE) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(Compositional)
library(MCMCpack)
library(gtools)
```



## Announcements

<!-- - Take Survey I. -->

- Homework 4 due tomorrow.

--

## Outline

- Categorical data

- Dirichlet distribution

- Conjugacy



---
## Categorical data (univariate)

- Suppose

  + $Y \in \{1,\ldots, d\}$;
  
  + $\Pr(Y = j) = \theta_j$ for each $j = 1,\ldots, d$; and
  
  + $\boldsymbol{\theta} = (\theta_1,\ldots,\theta_d)$.
  
- Then the pmf of $Y$ is
.block[
.small[
$$\Pr[Y = j| \boldsymbol{\theta}] = \prod_{j=1}^d \theta_j^{\mathbb{1}[Y = j]}.$$
]
]

--

- We say $Y$ has a .hlight[multinomial distribution] with sample size 1, or a .hlight[categorical distribution].

--

- Write as $Y | \boldsymbol{\theta} \sim \textrm{Multinomial}(1,\boldsymbol{\theta})$ or $Y | \boldsymbol{\theta} \sim \textrm{Categorical}(\boldsymbol{\theta})$.

--

- Clearly, this is just an extension of the Bernoulli distribution.




---
## Dirichlet distribution

- Since the elements of the probability vector $\boldsymbol{\theta}$ must always sum to one, the support is often called a .hlight[simplex].

--

- A conjugate prior for categorical/multinomial data is the .hlight[Dirichlet distribution].

--

- A random variable $\boldsymbol{\theta}$ has a .hlight[Dirichlet distribution] with parameter $\boldsymbol{\alpha}$, if
.block[
.small[
$$p[\boldsymbol{\theta} | \boldsymbol{\alpha}] = \dfrac{\Gamma\left(\sum_{j=1}^d \alpha_j\right)}{\prod_{j=1}^d \Gamma(\alpha_j)} \prod_{j=1}^d \theta_j^{\alpha_j-1}, \ \ \ \alpha_j > 0 \ \textrm{ for all } \ j = 1,\ldots, d.$$
]
]

  where $\boldsymbol{\alpha} = (\alpha_1,\ldots,\alpha_d)$, and
.block[
.small[
$$\sum_{j=1}^d \theta_j = 1, \ \ \theta_j \geq 0 \ \textrm{ for all } \ j = 1,\ldots, d.$$
]
]

--

- We write this as $\boldsymbol{\theta} \sim \textrm{Dirichlet}(\boldsymbol{\alpha}) = \textrm{Dirichlet}(\alpha_1,\ldots,\alpha_d)$.

--

- The Dirichlet distribution is a multivariate generalization of the .hlight[beta distribution].



---
## Dirichlet distribution

- Write
.block[
.small[
$$\alpha_0 = \sum_{j=1}^d \alpha_j \ \ \ \textrm{and} \ \ \ \alpha_j^\star = \dfrac{\alpha_j}{\alpha_0}.$$
]
]

--

- Then we can re-write the pdf slightly as
.block[
.small[
$$p[\boldsymbol{\theta} | \boldsymbol{\alpha}] = \dfrac{\Gamma\left(\alpha_0\right)}{\prod_{j=1}^d \Gamma(\alpha_j)} \prod_{j=1}^d \theta_j^{\alpha_j-1}, \ \ \ \alpha_j > 0 \ \textrm{ for all } \ j = 1,\ldots, d.$$
]
]
  
--

- Properties:

  - 
.block[
.small[
$$\mathbb{E}[\theta_j] = \alpha_j^\star;$$
]
]

--

  - 
.block[
.small[
$$\textrm{Mode}[\theta_j] = \dfrac{\alpha_j - 1}{\alpha_0 - d};$$
]
]

--
  
  - 
.block[
.small[
$$\mathbb{Var}[\theta_j] = \dfrac{\alpha_j^\star(1-\alpha_j^\star)}{\alpha_0 + 1} = \dfrac{\mathbb{E}[\theta_j](1-\mathbb{E}[\theta_j])}{\alpha_0 + 1};$$
]
]
  
--

  - 
.block[
.small[
$$\mathbb{Cov}[\theta_j,\theta_k] = \dfrac{\alpha_j^\star\alpha_k^\star}{\alpha_0 + 1} = \dfrac{\mathbb{E}[\theta_j]\mathbb{E}[\theta_k]}{\alpha_0 + 1}.$$
]
]





---
## Dirichlet examples

$\textrm{Dirichlet}(1,1,1)$

```{r fig.height=5, echo=F}
draws <- rdirichlet(1000, c(1,1,1))
bivt.contour(draws)
```



---
## Dirichlet examples

$\textrm{Dirichlet}(10,10,10)$

```{r fig.height=5, echo=F}
draws <- rdirichlet(1000, c(10,10,10))
bivt.contour(draws)
```



---
## Dirichlet examples

$\textrm{Dirichlet}(10,10,10)$

```{r fig.height=5, echo=F}
draws <- rdirichlet(1000, c(100,100,100))
bivt.contour(draws)
```



---
## Dirichlet examples

$\textrm{Dirichlet}(1,10,1)$

```{r fig.height=5, echo=F}
draws <- rdirichlet(1000, c(1,10,1))
bivt.contour(draws)
```




---
## Dirichlet examples

$\textrm{Dirichlet}(50,100,10)$

```{r fig.height=5, echo=F}
draws <- rdirichlet(1000, c(50,100,10))
bivt.contour(draws)
```





---
## Likelihood

- Let $Y_i, \ldots, Y_n | \boldsymbol{\theta} \sim \textrm{Categorical}(\boldsymbol{\theta})$.

--

- Recall
.block[
.small[
$$\Pr[Y_i = j| \boldsymbol{\theta}] = \prod_{j=1}^d \theta_j^{\mathbb{1}[Y_i = j]}.$$
]
]

--

- Then,
.block[
.small[
$$
\begin{split}
L[Y; \boldsymbol{\theta}] & = \prod_{i=1}^n \prod_{j=1}^d \theta_j^{\mathbb{1}[Y_i = j]} = \prod_{j=1}^d  \theta_j^{\sum_{i=1}^n \mathbb{1}[Y_i = j]} = \prod_{j=1}^d  \theta_j^{n_j}
\end{split}
$$
]
]

  where $n_j$ is just the number of individuals in category $j$.

--

- Maximum likelihood estimate of $\theta_j$ is
.block[
.small[
$$\hat{\theta}_j = \dfrac{n_j}{n}, \ \ j = 1,\ldots, d$$
]
]



---
## Posterior

- Set $\pi(\boldsymbol{\theta}) = \textrm{Dirichlet}(\alpha_1,\ldots,\alpha_d)$.
.block[
.small[
$$
\begin{split}
\pi(\boldsymbol{\theta} | Y) & \propto L[Y; \boldsymbol{\theta}] \pi[\boldsymbol{\theta}]\\
& \propto \prod_{j=1}^d  \theta_j^{n_j} \prod_{j=1}^d \theta_j^{\alpha_j-1} \\
& \propto \prod_{j=1}^d  \theta_j^{\alpha_j + n_j - 1}\\
& = \textrm{Dirichlet}(\alpha_1 + n_1,\ldots,\alpha_d + n_d)
\end{split}
$$
]
]

--

- Posterior expectation:
.block[
.small[
$$\mathbb{E}[\theta_j | Y] = \dfrac{\alpha_j + n_j}{\sum_{l=1}^d (\alpha_l + n_l)}.$$
]
]



---
## Combining information

- For the prior, we have
.block[
.small[
$$\mathbb{E}[\theta_j] = \dfrac{\alpha_j}{\sum_{j=1}^d \alpha_j}$$
]
]

--

- We can think of 

  + $\theta_{0j} = \mathbb{E}[\theta_j]$ as being our **"prior guess"** about $\theta_j$, and
  
  + $n_0 = \sum_{j=1}^d \alpha_j$ as being our **"prior sample size"**.
  
  

--

- We can then rewrite the prior as $\pi(\boldsymbol{\theta}) = \textrm{Dirichlet}(n_0\theta_{01},\ldots,n_0\theta_{0d})$.




---
## Combining information

- We can write the posterior expectation as:
.block[
.small[
$$
\begin{split}
\mathbb{E}[\theta_j | Y] & = \dfrac{\alpha_j + n_j}{\sum_{l=1}^d (\alpha_l + n_l)}\\
& = \dfrac{\alpha_j}{\sum_{l=1}^d \alpha_l + \sum_{l=1}^d n_l} + \dfrac{n_j}{\sum_{l=1}^d \alpha_l + \sum_{l=1}^d n_l}\\
& = \dfrac{n_0\theta_{0j}}{n_0 + n} + \dfrac{n \hat{\theta}_j}{n_0 + n}\\
& = \dfrac{n_0}{n_0 + n} \theta_{0j} + \dfrac{n}{n_0 + n} \hat{\theta}_j.
\end{split}
$$
]
]

  since
MLE is
.block[
.small[
$$\hat{\theta}_j = \dfrac{n_j}{n}$$
]
]

--

- Once again, we can express our posterior expectation as a weighted average of the prior expectation and MLE.

--

- <div class="question">
We can also extend the Dirichlet-multinomial model to more variables (contingency tables).
</div>




---
## Example: pre-election polling

- Fox News Nov 3-6 pre-election survey of 1295 likely voters for the 2016 election.

--

- For those interested, [FiveThirtyEight](https://projects.fivethirtyeight.com/) is an interesting source for pre-election polls.

--

- Out of 1295 respondents, 622 indicated support for Clinton, 570 for Trump, and the remaining 103 for other candidates or no opinion.

--

- Drawing inference from pre-election polls is way more complicated and nuanced that this. We only use the data here for this simple illustration.

--

- Assuming no other information on the respondents, we can assume simple random sampling and use a multinomial distribution with parameter $\boldsymbol{\theta} = (\theta_1,\theta_2,\theta_3)$, the proportion, in the survey population, of Clinton supporters, Trump supporters and other candidates or no opinion.



---
## Example: pre-election polling

- With a noninformative uniform prior, we have $\pi(\boldsymbol{\theta}) = \textrm{Dirichlet}(1,1,1)$.

--

- The resulting posterior is $\textrm{Dirichlet}(1 + n_1,1 + n_2,1 + n_3) = \textrm{Dirichlet}(623,571,104)$.

--

- Suppose we wish to compare the proportion of people who would vote for Trump versus Clinton, we could examine the posterior distribution of $\theta_1-\theta_2$.

--

- We can even compute the probability $\Pr(\theta_1 > \theta_2 | Y)$.


---
## Example: pre-election polling

```{r fig.height=3.8}
#library(gtools)
PostSamples <- rdirichlet(10000, alpha=c(623,571,104))
#dim(PostSamples)
hist((PostSamples[,1] - PostSamples[,2]),col=rainbow(20),xlab=expression(theta[1]-theta[2]),
     ylab="Posterior Density",freq=F,breaks=50,
     main=expression(paste("Posterior density of ",theta[1]-theta[2])))
```



---
## Example: pre-election polling

- Posterior probability that Clinton had more support than Trump in the survey population, that is, $\Pr(\theta_1 > \theta_2 | Y)$, is

  ```{r fig.height=3.5}
#library(gtools)
mean(PostSamples[,1] > PostSamples[,2])
```

--

- Once again, this is just a simple illustration with a very small subset of the 2016 pre-election polling data.

--

- Inference for pre-election polls is way more complex and nuanced that this.












